---
title: "Data Analysis"
subtitle: "Chapter 12: Other Aesthetics"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "`ggplot2`: Elegant Graphics for Data Analysis"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch12-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

# Other Aesthetics

## Overview

Position and colour are the two most prominent aesthetics in `ggplot2`, but a rich set of additional aesthetics is available for encoding data—especially when multiple variables must be represented simultaneously or when a colour-only approach is insufficient.

This chapter covers:

- **Size** — mapping continuous values to the area or radius of point markers
- **Shape** — mapping discrete values to point marker shapes
- **Line width** — controlling the width of lines independently of size
- **Line type** — mapping discrete values to solid, dashed, dotted, or custom dash patterns
- **Manual scales** — assigning values to aesthetics by explicit hand-crafted lookup tables
- **Identity scales** — passing pre-scaled data values directly through to the aesthetic without transformation

# Size

## Size

The `size` aesthetic is most commonly used to scale point markers and text labels. The default scale, `scale_size()`, maps a linear increase in the data variable to a linear increase in the **area** of the geom — not its radius. Area scaling is the perceptually justified default because human judgements of relative size are better calibrated to area than to linear dimension.

By default, the smallest value in the data is rendered at size 1 and the largest at size 6. The `range` argument overrides this:

---

```{r, echo=TRUE, warning=FALSE, message=FALSE, out.width="60%", fig.align='center'}
library(ggplot2)
library(patchwork)

base <- ggplot(mpg, aes(displ, hwy, size = cyl)) +
  geom_point()

p_def   <- base
p_small <- base + scale_size(range = c(1, 2))
p_def | p_small
```

---

A family of related size scales covers specialised situations:

| Scale | Description |
|---|---|
| `scale_size_area()` | Area scaling where data value 0 maps to area 0 |
| `scale_size_binned_area()` | Binned version of `scale_size_area()` |
| `scale_radius()` | Maps data values to point **radius** rather than area |
| `scale_size_binned()` | Continuous variable mapped to discrete size steps |
| `scale_size_date()` / `scale_size_datetime()` | Size scales for date/date-time variables |

---

### Radius Size Scales

Area scaling is not always appropriate. When the data variable is itself a physical radius or diameter, radius scaling should be used instead, via `scale_radius()`.

Consider this dataset of planetary radii:

---

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
planets <- data.frame(
  name     = c("Mercury","Venus","Earth","Mars",
               "Jupiter","Saturn","Uranus","Neptune"),
  type     = c(rep("Inner", 4), rep("Outer", 4)),
  position = 1:8,
  radius   = c(2440, 6052, 6378, 3390, 71400, 60330, 25559, 24764),
  orbit    = c(5.79e7, 1.08e8, 1.50e8, 2.28e8,
               7.78e8, 1.43e9, 2.87e9, 4.50e9)
)

base_pl <- ggplot(planets, aes(1, name, size = radius)) +
  geom_point() +
  scale_x_continuous(breaks = NULL) +
  labs(x = NULL, y = NULL, size = NULL)

base_pl + ggtitle("Default scale_size() — not to scale")
```

---

With the default `scale_size()`, the area of each dot scales with radius², distorting the apparent ratios. Using `scale_radius()` with `limits = c(0, NA)` anchors the zero-radius point at zero area and renders proportions faithfully:

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
base_pl +
  scale_radius(limits = c(0, NA), range = c(0, 10)) +
  ggtitle("scale_radius() — proportionally accurate")
```

The contrast is immediately visible: Jupiter's disc is now visibly much larger than Saturn's, correctly reflecting that the difference between the two exceeds the diameter of Earth.

---

### Binned Size Scales

`scale_size_binned()` discretises a continuous size variable into a fixed number of size steps, analogously to the binned colour scales seen in §11.4. The legend for a binned size scale is governed by `guide_bins()` rather than `guide_legend()`.

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
base_bin <- ggplot(mpg, aes(displ, manufacturer, size = hwy)) +
  geom_point(alpha = .2) +
  scale_size_binned()

base_bin
```

Unlike `guide_legend()`, which lays out keys in a table, `guide_bins()` arranges them along a single axis with an optional axis line.

---

Key arguments of `guide_bins()`:

| Argument | Effect |
|---|---|
| `axis` | Show/hide the axis line alongside the keys (default `TRUE`) |
| `direction` | `"vertical"` (default) or `"horizontal"` layout |
| `show.limits` | Whether tick marks appear at the scale endpoints (default `FALSE`) |
| `axis.colour` | Colour of the guide axis line |

---

| Argument | Effect |
|---|---|
| `axis.linewidth` | Width of the guide axis line |
| `axis.arrow` | Add an arrowhead to the axis using `arrow()` |
| `keywidth` / `keyheight` | Key dimensions (grid units) |
| `reverse` | Reverse the order of the keys |
| `override.aes` | Override geom aesthetics in the legend keys |

---

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
p_horiz <- base_bin +
  guides(size = guide_bins(direction = "horizontal"))

p_lim <- base_bin +
  guides(size = guide_bins(show.limits = TRUE))

p_horiz | p_lim
```

---

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
base_bin +
  guides(
    size = guide_bins(
      axis.arrow = arrow(
        length = unit(.1, "inches"),
        ends = "first",
        type = "closed"
      )
    )
  )
```

# Shape

## Shape

The `shape` aesthetic maps a discrete variable to the marker symbol used for each point. It is suited to variables with at most six distinct categories; beyond that, distinguishing shapes reliably becomes difficult and `ggplot2` will issue a warning.

The default `scale_shape()` provides a mixed palette of three solid and three hollow shapes when `solid = TRUE` (the default), or six hollow shapes when `solid = FALSE`:

---

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
base_shp <- ggplot(mpg, aes(displ, hwy, shape = factor(cyl))) +
  geom_point()

p_solid  <- base_shp
p_hollow <- base_shp + scale_shape(solid = FALSE)
p_solid | p_hollow
```

---

`ggplot2` supports 25 distinct shapes, each identified by an integer from 0 to 24:

- Shapes **0–14** are hollow (border only, no fill).
- Shapes **15–20** are solid (filled with the `colour` aesthetic).
- Shapes **21–24** are filled shapes that respond to both `colour` (border) and `fill` aesthetics independently.

To assign shapes manually, use `scale_shape_manual()` with a named integer vector:

---

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
base_shp +
  scale_shape_manual(
    values = c("4" = 16, "5" = 17, "6" = 1, "8" = 2)
  )
```

The names of the `values` vector correspond to the levels of the discrete variable; the integer values correspond to the desired shape codes.

# Line Width

## Line Width

The `linewidth` aesthetic, introduced in `ggplot2` 3.4.0, controls the width of rendered lines. Prior to this version, `size` served dual duty for both point size and line width, which caused ambiguity for compound geoms such as `geom_pointrange()` that combine both elements.

The three plots below illustrate the separation: the left uses default values for both aesthetics, the middle increases only `size` (enlarging the point endpoints), and the right increases only `linewidth` (thickening the connecting lines):

---

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
base_air <- ggplot(airquality, aes(x = factor(Month), y = Temp))

p_def <- base_air +
  geom_pointrange(stat = "summary", fun.data = "median_hilow")
p_sz  <- base_air +
  geom_pointrange(stat = "summary", fun.data = "median_hilow", size = 2)
p_lw  <- base_air +
  geom_pointrange(stat = "summary", fun.data = "median_hilow", linewidth = 2)

p_def | p_sz | p_lw
```

---

Although `linewidth` is most often set as a fixed constant, it is a true aesthetic and can be mapped to a data variable. Unlike `scale_size()`, where area increases proportionally to the data value, `scale_linewidth()` maps the data value to a **linear increase** in line width:

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
ggplot(airquality, aes(Day, Temp, group = Month)) +
  geom_line(aes(linewidth = Month)) +
  scale_linewidth(range = c(0.5, 3))
```

Binned linewidth scales are also available via `scale_linewidth_binned()`, following the same API as other binned scales.

# Line Type

## Line Type

The `linetype` aesthetic maps a discrete variable to the dash pattern of rendered lines. `scale_linetype()` is an alias for `scale_linetype_discrete()`. Continuous variables require `scale_linetype_binned()`; a `scale_linetype_continuous()` function exists but produces an error by design, as continuous line-type gradients have no perceptual basis.

Line type is best reserved for variables with a small number of categories. Even with five categories the plot can be difficult to parse:

---

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
ggplot(economics_long, aes(date, value01, linetype = variable)) +
  geom_line()
```

---

The default palette is `scales::linetype_pal()`, which provides 13 distinct line patterns. To see all available types:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
df_lt <- data.frame(value = letters[1:13])
ggplot(df_lt, aes(linetype = value)) +
  geom_segment(
    mapping = aes(x = 0, xend = 1, y = value, yend = value),
    show.legend = FALSE
  ) +
  theme(panel.grid = element_blank()) +
  scale_x_continuous(NULL, NULL)
```

---

### Custom Line Types

Line patterns can also be specified as strings of up to **8 hexadecimal characters** (0–F). Each character encodes the length of alternating drawn-segment and gap intervals. For example, `"55"` means: draw 5 units, gap 5 units (a regular dashed line).

Custom line types can be applied in two ways:

1. Using `scale_linetype_manual()` with a named character vector of hex strings.
2. Passing a custom palette function to the `palette` argument of `discrete_scale()`.

---

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
linetypes <- function(n) {
  types <- c("55", "75", "95", "1115", "111115",
             "11111115", "5158", "9198", "c1c8")
  types[seq_len(n)]
}

df_lt2 <- data.frame(value = letters[1:9])
ggplot(df_lt2, aes(linetype = value)) +
  geom_segment(
    mapping = aes(x = 0, xend = 1, y = value, yend = value),
    show.legend = FALSE
  ) +
  theme(panel.grid = element_blank()) +
  scale_x_continuous(NULL, NULL) +
  discrete_scale("linetype", palette = linetypes)
```

---

When the number of categories exceeds the number of patterns the palette defines, the excess levels receive `NA`, which by default renders as a blank (invisible) line. Override this with the `na.value` argument:

```{r, echo=TRUE, eval=FALSE}
discrete_scale("linetype", palette = linetypes, na.value = "dotted")
```

Human-readable string aliases are also valid linetype specifications: `"blank"`, `"solid"`, `"dashed"`, `"dotted"`, `"dotdash"`, `"longdash"`, and `"twodash"`.

# Manual Scales

## Manual Scales

Manual scales provide an explicit lookup table that maps each discrete data value to a specific aesthetic value. Every aesthetic in `ggplot2` has a corresponding manual scale: `scale_colour_manual()`, `scale_fill_manual()`, `scale_shape_manual()`, `scale_linetype_manual()`, `scale_size_manual()`, and so on.

The central argument is `values`: a vector (optionally named) of aesthetic values. When named, ggplot2 matches by name; when unnamed, it matches in the order of the factor levels.

Valid aesthetic values for each aesthetic type are documented in `vignette("ggplot2-specs")`.

---

### Constructing Legends with Manual Colour Scales

A particularly powerful application of `scale_colour_manual()` is creating an informative legend when multiple variables are displayed on the same plot. Consider this pair of time series offset by ±5 units from the `LakeHuron` water level:

---

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
huron <- data.frame(year = 1875:1972, level = as.numeric(LakeHuron))

# Naïve approach: colour set as a fixed parameter — no legend possible
ggplot(huron, aes(year)) +
  geom_line(aes(y = level + 5), colour = "red") +
  geom_line(aes(y = level - 5), colour = "blue")
```

When colour is set as a fixed constant outside `aes()`, `ggplot2` has no way to generate a legend for it.

---

The solution is to **map a label string to the colour aesthetic** inside `aes()`, and then instruct the manual scale how to translate those labels into actual colours:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
# Step 1: map a descriptive label to colour inside aes()
ggplot(huron, aes(year)) +
  geom_line(aes(y = level + 5, colour = "above")) +
  geom_line(aes(y = level - 5, colour = "below"))
```

ggplot2 now recognises `colour` as an aesthetic mapping and automatically generates a legend — though the default colours are not meaningful yet.

---

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
# Step 2: map label names to intended colours via scale_colour_manual()
ggplot(huron, aes(year)) +
  geom_line(aes(y = level + 5, colour = "above")) +
  geom_line(aes(y = level - 5, colour = "below")) +
  scale_colour_manual(
    name   = "Direction",
    values = c("above" = "red", "below" = "blue")
  )
```

The result is a fully functional legend without requiring a reshaped ("tidy long") dataset. This pattern is general: any time you need a legend for what would otherwise be a fixed aesthetic, map a character label to the aesthetic inside `aes()` and decode it through the corresponding manual scale.

# Identity Scales

## Identity Scales

Identity scales pass data values directly to the aesthetic **without any transformation**. They are appropriate when the data column already contains valid aesthetic values — for example, a column of colour names such as `"red"` or hex codes such as `"#FF0000"`, or a column of integer shape codes.

The available identity scales include `scale_colour_identity()`, `scale_fill_identity()`, `scale_shape_identity()`, `scale_linetype_identity()`, and `scale_size_identity()`.

---

A compelling example is the `luv_colours` dataset, which records the locations of all R built-in colour names in the LUV perceptual colour space (the space on which HCL is based):

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
head(luv_colours)
```

Each row contains a `col` column holding the actual R colour name for that point. Mapping `col` to the `colour` aesthetic and applying `scale_color_identity()` renders each point in its own colour — the data and aesthetic spaces are identical, so no legend is needed or meaningful:

---

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
ggplot(luv_colours, aes(u, v)) +
  geom_point(aes(colour = col), size = 3) +
  scale_color_identity() +
  coord_equal()
```

The plot is itself the legend: each dot's colour is the colour it represents. No additional key is required.

---

**When to use identity scales**

Identity scales are appropriate in two situations:

1. **Pre-scaled data**: the dataset was constructed with colour names, shape codes, or other aesthetic values already embedded — common when combining outputs from statistical models or external tools.

2. **Programmatic colour assignment**: when colour choices are computed algorithmically (e.g., by a custom clustering function) and you want `ggplot2` to honour them verbatim rather than applying its own mapping.

In all other situations, a properly specified aesthetic mapping with an appropriate scale is preferred, because it enables legend generation and ensures reproducible, principled colour choices.

---

**Exercises**

1. Using the `mpg` dataset, create a scatterplot of `displ` vs `hwy` with `size = cyl`. Compare the default `scale_size()` with `scale_size_area()`. When would you prefer the latter?

2. Build a bubble chart of the `planets` data showing each planet's orbital distance on the x-axis and its name on the y-axis, with bubble size proportional to planetary radius. Show why `scale_radius(limits = c(0, NA))` is necessary for an honest visual comparison.

3. Create a scatterplot using `mpg` with `shape = class`. Why does ggplot2 warn when `class` has more than 6 levels? How does `scale_shape_manual()` let you work around this?

---

4. Using `airquality`, build a `geom_pointrange()` plot of monthly temperature summaries. Demonstrate the independent effect of increasing `size` vs `linewidth` on the appearance of the geom.

5. Plot `economics_long` with `linetype = variable`. Apply `scale_linetype_manual()` to assign custom hex-pattern line types to each variable. What are the readability limits of the linetype aesthetic?

6. Recreate the `LakeHuron` example from §12.5 using `scale_linetype_manual()` instead of colour to distinguish the two series. Which approach (colour or line type) communicates the grouping more clearly, and why?

7. Using `luv_colours`, reproduce the identity-scale scatterplot. Then add a second layer using only the subset of colours whose names contain `"blue"`, drawn as larger points with a black border (`shape = 21`). What does `scale_fill_identity()` do in this context?

## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from [**ggplot2: Elegant Graphics for Data Analysis (3e)**](https://ggplot2-book.org/scales-other.html). All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.

