---
title: "Data Analysis"
subtitle: "Chapter 14: Scales and Guides"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "`ggplot2`: Elegant Graphics for Data Analysis"
logo: "logo.svg"
editor: source
bibliography: bib.bib
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch14-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

# Scales and Guides

## Overview

The scales toolbox in Chapters 10–12 solved practical visualisation problems. This chapter steps back to the **theory** that underpins scales and guides, showing that ideas introduced for position or colour scales apply across *every* aesthetic.

This chapter covers:

- **Theory** — scales as functions from data space to aesthetic space, and guides as their inverse
- **Scale names** — axis labels and legend titles as the same object
- **Breaks and limits** — the values a guide displays, and out-of-bounds handling
- **Scale guides** — the default guide for each scale type
- **Transformations** — transforming non-position aesthetics
- **Legend merging and splitting** — and customising legend key glyphs

# Theory of Scales and Guides

## Scales and Guides

Formally, a **scale** is a function from a region of *data space* (its domain) to a region of *aesthetic space* (its range). The **guide** — an axis or a legend — is the inverse function: it lets you read visual properties back to data values.

It may surprise you that axes and legends are the *same kind of thing*. They look very different but share one purpose, and their components line up:

| Argument | Axis | Legend |
|---|---|---|
| `name` | Label | Title |
| `breaks` | Ticks & grid lines | Keys |
| `labels` | Tick labels | Key labels |

---

Legends are more complex than axes in three ways:

1. A legend can display **multiple aesthetics** (e.g. colour and shape) from **multiple layers**, and the key symbol depends on the geom used.
2. Axes always appear in the same place; **legends can be positioned** anywhere, so they need a global positioning system.
3. Legends have many tweakable details: orientation, number of columns, key size.

---

### Scale Specification

Every aesthetic is associated with **exactly one** scale. When you write a mapping, ggplot2 silently adds a default scale for each aesthetic:

```{r, echo=TRUE, warning=FALSE, message=FALSE, eval=FALSE}
library(ggplot2)
library(patchwork)
# What you write:
ggplot(mpg, aes(displ, hwy)) + geom_point(aes(colour = class))

# What ggplot2 does behind the scenes:
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

The default depends on the aesthetic and the variable type: a continuous `hwy` on `y` yields `scale_y_continuous()`; a discrete `class` on colour yields `scale_colour_discrete()`.

---

Adding a scale with `+` actually **overrides** any existing scale for that aesthetic — the last one wins. ggplot2 warns you when this happens:

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
library(ggplot2)
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_sqrt() +
  scale_colour_brewer()
```

Here `scale_x_sqrt()` replaces the default x scale and `scale_colour_brewer()` replaces the default colour scale.

---

### Naming Scheme & Fundamental Types

User-facing scale functions follow a three-part naming scheme separated by `_`:

1. `scale`
2. the primary **aesthetic** (`colour`, `shape`, `x`, …)
3. the **name** of the scale (`continuous`, `discrete`, `brewer`, …)

Internally, **all** scales belong to one of three fundamental types, each built by a constructor you should never need to call directly:

| Fundamental type | Constructor |
|---|---|
| Continuous | `continuous_scale()` |
| Discrete | `discrete_scale()` |
| Binned | `binned_scale()` |

# Scale Limits

## Scale Limits

Scale **limits** define the *region* of data space over which the mapping is defined. The region differs by fundamental type: continuous and binned scales are defined by two end points, while discrete scales must **enumerate** their categories.

This raises a question: what should ggplot2 do with **out-of-bounds** values that fall outside the limits? The behaviour is governed by the `oob` argument:

- `scales::oob_censor()` (default) — replace out-of-bounds values with `NA`
- `scales::oob_squish()` — squish out-of-bounds values into the range

---

```{r, echo=TRUE, warning=FALSE, out.width="32%", fig.align='center'}
library(ggplot2)
df <- data.frame(x = 1:6, y = 8:13)
base <- ggplot(df, aes(x, y)) +
  geom_col(aes(fill = x)) +
  geom_vline(xintercept = 3.5, colour = "red")

a <- base
b <- base + scale_fill_gradient(limits = c(1, 3))
c <- base + scale_fill_gradient(limits = c(1, 3), oob = scales::squish)
```

---

```{r, echo=TRUE, warning=FALSE, out.width="32%", fig.align='center'}
library(patchwork)
a|b|c
```

Left: default fill. Middle: the three rightmost bars fall outside `limits = c(1, 3)` and become grey `NA`. Right: `oob = scales::squish` clamps them into range instead.

# Scale Guides

## Scale Guides

Where `name` takes text, the `guide` argument takes a **guide object** created by a guide function. Each scale type has a default guide:

| Scale type | Default guide |
|---|---|
| continuous colour/fill | `guide_colourbar()` |
| binned colour/fill | `guide_coloursteps()` |
| position scales | `guide_axis()` |
| discrete (non-position) | `guide_legend()` |
| binned (other) | `guide_bins()` |

Each guide function carries many arguments — equivalent to theme settings (text colour, size, font) but scoped to a single guide. Those settings are covered in Chapter 17.

# Scale Transformation

## Scale Transformation

Scale transformations are most common on continuous **position** scales (§10.1.7), but they apply to other aesthetics too. Often the purpose is visual emphasis. For the Old Faithful density, a square-root fill transform reveals the low-density regions around each peak:

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
library(ggplot2)
base <- ggplot(faithfuld, aes(waiting, eruptions)) +
  geom_raster(aes(fill = density)) +
  scale_x_continuous(NULL, NULL, expand = c(0, 0)) +
  scale_y_continuous(NULL, NULL, expand = c(0, 0))
a <- base
b <- base + scale_fill_continuous(trans = "sqrt")
```

---

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
library(patchwork)
(a|b)
```

---

Transforming a **size** aesthetic changes its interpretation. On the left, `z` reads as a *weight* (larger = bigger group); reversing the scale makes `z` read as a *distance* (larger = appears smaller):

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
set.seed(1)
df <- data.frame(x = runif(20), y = runif(20), z = sample(20))
base <- ggplot(df, aes(x, y, size = z)) + geom_point()
a <- base
b <- base + scale_size(trans = "reverse")
```

---

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
library(patchwork)
(a|b)
```


# Legend Merging and Splitting

## Merging Legends

Position scales correspond one-to-one with axes, but the link between non-position scales and legends is looser: one legend may draw from multiple layers (**merging**), or one aesthetic may need multiple legends (**splitting**).

ggplot2 uses the **fewest** legends possible, combining those where the same variable is mapped to different aesthetics. If both `colour` and `shape` map to the same variable, a single legend results:

```{r, echo=TRUE, warning=FALSE, out.width="32%", fig.align='center'}
library(ggplot2)
toy <- data.frame(const = 1, up = 1:4, txt = letters[1:4])
base <- ggplot(toy, aes(const, up)) + scale_x_continuous(NULL, breaks = NULL)
a <- base + geom_point(aes(colour = txt))
b <- base + geom_point(aes(shape = txt))
c <- base + geom_point(aes(shape = txt, colour = txt))
```

---

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
library(patchwork)
(a|b|c)
```

---

For legends to merge they must share the same `name`. If you rename one scale, rename them all — `labs()` is a convenient way to do so:

```{r, echo=TRUE, warning=FALSE, out.width="40%", fig.align='center'}
base <- ggplot(toy, aes(const, up)) +
  geom_point(aes(shape = txt, colour = txt)) +
  scale_x_continuous(NULL, breaks = NULL)
a <- base + labs(shape = "Split legend")                              # two legends
b <- base + labs(shape = "Merged legend", colour = "Merged legend")   # one legend
```

---

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
library(patchwork)
(a|b)
```

You can also force or suppress a layer's appearance with the `show.legend` argument (`TRUE`/`FALSE`).

---

### Splitting Legends

Splitting is much rarer — mapping one aesthetic to two variables is usually inadvisable, so ggplot2 disallows it by default. The **ggnewscale** package overrides this: `new_scale_colour()` initialises a fresh colour scale. Commands above it apply to the first scale; commands below it apply to the second.

```{r, echo=TRUE, warning=FALSE, eval=FALSE}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = factor(year)), size = 5) +
  scale_colour_brewer("year", type = "qual", palette = 5)

base +
  ggnewscale::new_scale_colour() +
  geom_point(aes(colour = cyl == 4), size = 1, fill = NA) +
  scale_colour_manual("4 cylinder", values = c("grey60", "black"))
```

# Legend Key Glyphs

## Legend Key Glyphs

By default the glyph in a legend key matches the layer and aesthetic — coloured lines show up as coloured line segments, boxplots as small boxplots. To override this, use the `key_glyph` argument:

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
library(ggplot2)
base <- ggplot(economics, aes(date, psavert, colour = "savings"))
a <- base + geom_line()
b <- base + geom_line(key_glyph = "timeseries")
```

---

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
library(patchwork)
(a|b)
```

Each geom has an associated key-drawing function such as `draw_key_path()` or `draw_key_timeseries()`. You may pass the function directly: `key_glyph = draw_key_timeseries`.

---

**Exercises**

1. Take `ggplot(mpg, aes(displ, hwy)) + geom_point(aes(colour = class))` and write out the three default scales that ggplot2 adds for you. Why is specifying them manually tedious?

2. Build a `geom_col()` chart with `aes(fill = x)` and `scale_fill_gradient(limits = c(1, 3))`. Compare the default `oob` with `scales::squish`. When is each appropriate?

3. List the default guide for each fundamental scale type. Which guide is shared by continuous, binned, and discrete position scales?

---

4. Reproduce the Old Faithful `faithfuld` raster with and without `trans = "sqrt"` on the fill scale. What does the transformation make easier to see?

5. Using the `toy` data, create one plot where `colour` and `shape` map to the same variable and one where they map to differently named scales. Explain when legends merge.

6. Use `ggnewscale::new_scale_colour()` to overlay two `geom_point()` layers on `mpg` with separate colour scales. Why does ggplot2 disallow splitting by default?

7. Compare `geom_line()` with the default glyph against `key_glyph = "timeseries"` on the `economics` data. What other key-drawing functions can you find?

## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from [**ggplot2: Elegant Graphics for Data Analysis (3e)**](https://ggplot2-book.org/scales-guides.html). All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.

## References
