---
title: "Data Analysis"
subtitle: "Chapter 7: Networks"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "`ggplot2`: Elegant Graphics for Data Analysis"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch7-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

# Networks

## Networks

Like maps, network data occupies a specialised corner of the data visualisation landscape. While spatial data primarily requires handling map projections, network data introduces an entirely distinct data structure *and* an entirely distinct family of visual idioms.

ggplot2 does not natively support network visualisation. This chapter demonstrates how to use the **ggraph** package (<https://ggraph.data-imaginist.com>) — the most comprehensive and ggplot2-compatible solution available. Other packages offering related functionality include:

- `geomnet`, `ggnetwork`, and `GGally` — for general network plots.
- `ggtree` and `ggdendro` — specifically for tree and dendrogram visualisation.

------------------------------------------------------------------------

**Chapter roadmap**

- **Section 7.1** — the structure of network data and the `tidygraph` manipulation API.
- **Section 7.2** — visualising networks with `ggraph`: layouts, node geoms, edge geoms, and faceting.
- **Section 7.3** — pointers to further resources.

# What is network data?

## What is network data?

In mathematical terms, a **graph** (or **network**) is a data structure composed of two elements:

- **Nodes** (also called *vertices*) — the entities being modelled (people, genes, web pages, …).
- **Edges** (also called *links*) — the relationships connecting pairs of nodes (friendship, co-expression, hyperlinks, …).

Both nodes and edges may carry additional attributes. Edges may additionally be classified as:

- **Directed** — the relationship has a defined source and target (e.g., "child-of" is not symmetric).
- **Undirected** — the relationship is symmetric (e.g., mutual friendship).

------------------------------------------------------------------------

**Why network data is challenging for ggplot2**

Network data cannot be naturally represented as a single rectangular data frame, which is the fundamental assumption underlying ggplot2. It requires at minimum *two* interrelated tables — one for nodes and one for edges. The `tidygraph` package addresses this by providing a dual-table structure together with a dplyr-compatible API. `ggraph` builds on `tidygraph` to deliver a complete grammar of graphics for networks.

------------------------------------------------------------------------

## A tidy network manipulation API

`tidygraph` is best understood as a **dplyr interface for network data**. It exposes the same verbs (`mutate()`, `filter()`, `arrange()`, etc.) that analysts already know, but applies them to either the node table or the edge table of a graph.

The example below constructs a random directed graph using the Erdős–Rényi sampling model (`play_gnp()`), assigns a random categorical label to each node, and sorts the edges by the label of their source node:

---

```{r,echo=T,warning=FALSE,message=FALSE,out.width="80%",fig.align='center'}
library(tidygraph)
library(ggraph)
library(ggplot2)
library(dplyr)

set.seed(42)
graph <- play_gnp(n = 10, p = 0.2) |>
  activate(nodes) |>
  mutate(class = sample(letters[1:4], n(), replace = TRUE)) |>
  activate(edges) |>
  arrange(.N()$class[from])

graph
```

------------------------------------------------------------------------

**Key functions introduced**

- `activate(nodes)` / `activate(edges)` — switches the "active" context so that subsequent dplyr verbs operate on the node table or the edge table respectively. Only one table can be active at a time.

- `.N()` — within an edge-context `mutate()` or `filter()`, `.N()` gives read access to the *node* data frame of the current graph. This allows edge attributes to reference node attributes (e.g., sorting edges by a property of their source node).

- `.E()` — gives read access to the *edge* data frame when working in a node context.

- `.G()` — gives access to the entire graph object.

------------------------------------------------------------------------

## Conversion

Network data arrives in many formats. `tidygraph` provides `as_tbl_graph()` to convert the most common R network objects into the unified `tbl_graph` format.

**Converting an edge-list data frame:**

The `highschool` dataset from `ggraph` records friendship nominations among high-school boys in two academic years (1957 and 1958), stored as a three-column edge list:

```{r,echo=T,warning=FALSE,message=FALSE}
data(highschool, package = "ggraph")
head(highschool)
```

---

```{r,echo=T,warning=FALSE,message=FALSE}
hs_graph <- as_tbl_graph(highschool, directed = FALSE)
hs_graph
```

------------------------------------------------------------------------

**Converting a hierarchical clustering result:**

`as_tbl_graph()` also handles the output of `hclust()`, automatically extracting node-level attributes such as `label`, `leaf` status, and linkage `height`:

```{r,echo=T,warning=FALSE,message=FALSE}
luv_clust <- hclust(dist(luv_colours[, 1:3]))
luv_graph <- as_tbl_graph(luv_clust)
luv_graph
```

---

When converting, `tidygraph` preserves any additional columns present in the input data (e.g., the `year` column from `highschool`, and the `label` and `leaf` columns from the clustering result) and attaches them as node or edge attributes automatically.

------------------------------------------------------------------------

## Algorithms

One of `tidygraph`'s most powerful features is its integrated library of **graph algorithms**. These include:

- **Centrality** — measures of node importance (e.g., degree, betweenness, PageRank, eigenvector).
- **Ranking** — orderings that place well-connected nodes near one another.
- **Grouping / community detection** — partitioning nodes into clusters based on edge density.

All algorithm functions are designed to be called **inside `mutate()`**. They return a vector whose length and ordering match the active node or edge table, and they infer the graph context implicitly — no need to pass the graph or node indices as arguments.

------------------------------------------------------------------------

**Example: PageRank centrality**

The code below adds a `centrality` column to the node table of `graph`, computed using the PageRank algorithm, and sorts nodes in descending order of centrality:

```{r,echo=T,warning=FALSE,message=FALSE}
graph |>
  activate(nodes) |>
  mutate(centrality = centrality_pagerank()) |>
  arrange(desc(centrality))
```

---

Because algorithms are evaluated lazily inside `mutate()`, they can also be invoked directly inside `aes()` calls in `ggraph` — enabling on-the-fly computation during plotting without modifying the underlying graph object.

------------------------------------------------------------------------

**Further resources for `tidygraph`**

This section provides only a brief orientation to `tidygraph`. A comprehensive overview of all available functions and algorithms is available at <https://tidygraph.data-imaginist.com>.

# Visualizing networks

## Visualizing networks

`ggraph` extends the grammar of graphics to network data by sitting on top of both `tidygraph` and `ggplot2`. While its API is deliberately familiar, it differs from standard ggplot2 usage in one important respect: most network visualisations do **not** map variables directly to `x` and `y` aesthetics.

Instead, node positions are determined by a **layout algorithm** that uses the graph's topology to assign (often mathematically arbitrary) `x` and `y` coordinates. The positions communicate structural relationships — connectivity, hierarchy, clustering — rather than the values of any measured variable.

------------------------------------------------------------------------

## Setting up the visualization

A ggraph plot is initialised with `ggraph()` rather than `ggplot()`. The function signature is:

```{r,echo=T,eval=FALSE}
ggraph(graph, layout = "...", ...)
```

- **`graph`** — a `tbl_graph`, or any object that `as_tbl_graph()` can convert.
- **`layout`** — either a string naming a built-in layout, or a custom function that accepts a `tbl_graph` and returns a data frame with at least `x` and `y` columns (one row per node).
- **`...`** — additional arguments forwarded to the layout function.

If `layout` is omitted, `ggraph` selects a sensible default based on the graph type.

------------------------------------------------------------------------

**Specifying a layout**

The three examples below illustrate: (1) using the default layout, (2) specifying a named layout explicitly, and (3) passing edge weights as a layout argument:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
# Default layout (stress)
ggraph(hs_graph) +
  geom_edge_link() +
  geom_node_point()
```

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
# Explicit layout: DrL force-directed
ggraph(hs_graph, layout = "drl") +
  geom_edge_link() +
  geom_node_point()
```

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
# Layout with edge weights; edge alpha encodes weight
hs_graph <- hs_graph |>
  activate(edges) |>
  mutate(edge_weights = runif(n()))

ggraph(hs_graph, layout = "stress", weights = edge_weights) +
  geom_edge_link(aes(alpha = edge_weights)) +
  geom_node_point() +
  scale_edge_alpha_identity()
```

------------------------------------------------------------------------

**A caution on layout choice**

Network layouts are well-known for their ability to create the visual impression of structure — clusters, hierarchies, or dense hubs — that may not reflect true patterns in the data. The same graph can look dramatically different under different layouts. Always explore multiple layouts before settling on one for publication, and be sceptical of apparent patterns that only appear in a single layout.

------------------------------------------------------------------------

**Circularity**

Some layouts support a `circular = TRUE` argument that arranges nodes in a ring rather than a flat plane. The correct way to achieve this in ggraph is via the layout parameter, **not** via `coord_polar()`. Using `coord_polar()` bends the *edges* as well as repositioning the nodes, which is almost never desirable:

---

```{r,echo=T,warning=FALSE,message=FALSE,out.width="65%",fig.align='center'}
library(patchwork)

p1 <- ggraph(luv_graph, layout = "dendrogram", circular = TRUE) +
  geom_edge_link() +
  coord_fixed() +
  ggtitle("circular = TRUE (correct)")

p2 <- ggraph(luv_graph, layout = "dendrogram") +
  geom_edge_link() +
  coord_polar() +
  scale_y_reverse() +
  ggtitle("coord_polar() (edges bent — avoid)")

p1 | p2
```

------------------------------------------------------------------------

## Drawing nodes

All node-drawing geoms in ggraph share the `geom_node_` prefix. The workhorse is `geom_node_point()`, which renders each node as a point — conceptually similar to `geom_point()`, but with three important differences shared by all node and edge geoms:

1. **`x` and `y` are implicit** — node positions are set by the layout and do not need to be specified in `aes()`.
2. **`filter` aesthetic** — a logical aesthetic that suppresses the drawing of specific nodes when set to `FALSE`, without removing them from the graph structure used by edges.
3. **In-`aes()` algorithm calls** — any `tidygraph` algorithm can be called directly inside `aes()`, and it will be evaluated on the graph being visualised. This enables rapid iteration without modifying the underlying data.

------------------------------------------------------------------------

**Example: filtering and colouring nodes by centrality**

The plot below renders the `hs_graph`, showing only nodes with more than 2 connections (filtered using `centrality_degree()`), and colours visible nodes by their power centrality:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
ggraph(hs_graph, layout = "stress") +
  geom_edge_link() +
  geom_node_point(
    aes(filter = centrality_degree() > 2,
        colour = centrality_power()),
    size = 4
  )
```

------------------------------------------------------------------------

**Specialised node geoms**

Beyond `geom_node_point()`, ggraph provides layout-specific node geoms. For example, `geom_node_tile()` is used when the chosen layout is `"treemap"` — in this context, each node is rendered as a rectangular tile whose area is proportional to its weight in the hierarchy:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
ggraph(luv_graph, layout = "treemap") +
  geom_node_tile(aes(fill = depth))
```

------------------------------------------------------------------------

## Drawing edges

Edge geoms share the `geom_edge_` prefix and offer considerably more variety than node geoms, reflecting the many ways in which a connection between two entities can be rendered.

----

**Three variants of every edge geom**

Every edge geom type in ggraph exists in three variants, distinguished by a numeric suffix:

| Variant | Suffix | Behaviour |
|---------|--------|-----------|
| Standard | (none) | Splits each edge into sub-segments; supports gradients along the edge |
| Simple | `0` | Draws the edge as a single primitive; faster but no gradient support |
| Interpolated | `2` | Interpolates between values at the two endpoints; accesses terminal node attributes via `node.` prefix |

------------------------------------------------------------------------

**`geom_edge_link()` — straight-line edges**

`geom_edge_link()` is the most commonly used edge geom. In its standard form, the edge is subdivided internally, enabling a gradient along its length. Mapping `after_stat(index)` to `alpha` produces a directional fade from source (transparent) to target (opaque):

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
ggraph(graph, layout = "stress") +
  geom_edge_link(aes(alpha = after_stat(index)))
```

------------------------------------------------------------------------

**`geom_edge_link2()` — interpolating node attributes**

The `2` variant allows the edge colour (or other aesthetic) to be interpolated between the values of an attribute at the two terminal nodes. Node attributes are accessed inside `aes()` with the `node.` prefix:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
ggraph(graph, layout = "stress") +
  geom_edge_link2(
    aes(colour = node.class),
    width = 3,
    lineend = "round"
  )
```

The `node1.` and `node2.` prefixes (available in the standard and `0` variants) give access to the source and target node attributes separately, while `node.` (used in the `2` variant) refers generically to the current endpoint being interpolated.

------------------------------------------------------------------------

**Handling parallel edges**

When multiple edges connect the same pair of nodes (a *multigraph*), drawing them as straight lines stacks them on top of one another and hides the multiplicity. Two geoms address this:

- `geom_edge_fan()` — spreads parallel edges into arcs that fan out between the two nodes.
- `geom_edge_parallel()` — offsets parallel edges as parallel straight lines.

---

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
p1 <- ggraph(hs_graph, layout = "stress") +
  geom_edge_fan() + ggtitle("geom_edge_fan()")

p2 <- ggraph(hs_graph, layout = "stress") +
  geom_edge_parallel() + ggtitle("geom_edge_parallel()")

p1 | p2
```

Both geoms increase visual clutter and are best reserved for relatively small graphs.

------------------------------------------------------------------------

**Tree and dendrogram edges**

For hierarchical layouts such as dendrograms, the conventional visual style uses right-angle connectors rather than straight diagonal lines. `geom_edge_elbow()` provides this:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
ggraph(luv_graph, layout = "dendrogram", height = height) +
  geom_edge_elbow()
```

Smoother alternatives that produce curved connectors are `geom_edge_bend()` and `geom_edge_diagonal()`.

------------------------------------------------------------------------

**Clipping edges around nodes**

When directional arrows are added to edges, a common problem arises: the arrowhead is hidden beneath the node point because the edge is drawn all the way to the node's geometric centre. The `start_cap` and `end_cap` aesthetics solve this by defining a **clipping region** around each terminal node — the edge is terminated before it enters the clipping zone.

*Without clipping (arrowhead obscured):*

```{r,echo=T,warning=FALSE,message=FALSE,out.width="65%",fig.align='center'}
ggraph(graph, layout = "stress") +
  geom_edge_link(arrow = arrow()) +
  geom_node_point(aes(colour = class), size = 8)
```

------------------------------------------------------------------------

*With clipping (arrowhead visible):*

```{r,echo=T,warning=FALSE,message=FALSE,out.width="65%",fig.align='center'}
ggraph(graph, layout = "stress") +
  geom_edge_link(
    arrow     = arrow(),
    start_cap = circle(5, "mm"),
    end_cap   = circle(5, "mm")
  ) +
  geom_node_point(aes(colour = class), size = 8)
```

`circle(5, "mm")` defines a circular clipping region of 5 mm radius centred on each terminal node. The `circle()` helper function is provided by ggraph; other geometry helpers (e.g., `square()`, `ellipsis()`) are also available.

------------------------------------------------------------------------

**Edges as non-line objects**

Edges do not have to be rendered as lines. In a **matrix plot** (also called an adjacency matrix plot), nodes are placed implicitly along both axes, and edges are shown as tiles or points at their row–column intersection. `geom_edge_point()` implements this for the matrix layout:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
ggraph(hs_graph, layout = "matrix", sort.by = node_rank_traveller()) +
  geom_edge_point()
```

Matrix plots scale to much larger graphs than node–link diagrams and are particularly effective for revealing block-diagonal structure in ordered adjacency matrices.

------------------------------------------------------------------------

## Faceting

Faceting is conceptually as valuable for networks as it is for tabular data — and for much the same reasons: it enables direct visual comparison of the same structure across subsets. However, standard `facet_wrap()` and `facet_grid()` cannot be used directly with ggraph because they do not understand the dependency between node and edge tables.

ggraph provides three specialised faceting functions:

- `facet_edges(~ var)` — splits edges into panels by a variable in the edge table; **nodes are repeated in all panels**.
- `facet_nodes(~ var)` — splits nodes into panels by a variable in the node table; **edges whose endpoints are in different panels are dropped**.
- `facet_graph(row_var ~ col_var, rows = "edges"/"nodes", cols = "edges"/"nodes")` — a two-dimensional equivalent of `facet_grid()` that allows independent specification of which table drives the rows and which drives the columns.

------------------------------------------------------------------------

**`facet_edges()` example: friendship evolution over time**

The `highschool` data records friendships in two separate years. Using `facet_edges(~year)` splits the edge table by year while repeating all 70 nodes in each panel, making the structural evolution directly comparable:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="75%",fig.align='center'}
ggraph(hs_graph, layout = "stress") +
  geom_edge_link() +
  geom_node_point() +
  facet_edges(~year)
```

The faceted view makes it immediately apparent that the network transitions from two near-completely disconnected groups in 1957 to a more integrated, single-component structure in 1958.

------------------------------------------------------------------------

**`facet_nodes()` example: community detection**

Because faceting accepts `tidygraph` algorithm calls directly, it is straightforward to visualise the result of a community detection algorithm without pre-computing and storing the partition:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="75%",fig.align='center'}
ggraph(hs_graph, layout = "stress") +
  geom_edge_link() +
  geom_node_point() +
  facet_nodes(~ group_spinglass())
```

`group_spinglass()` partitions the nodes using the spin-glass community detection algorithm. Each panel displays the subgraph induced by one detected community; edges between communities are dropped. The result allows immediate assessment of whether the detected communities are genuinely well-separated.

# Want more?

## Want more?

This chapter provides a practical introduction to network visualisation with `ggraph` and `tidygraph`. Many additional capabilities — including more layout algorithms, further edge and node geom types, legend handling for edges, and custom scale functions — are documented in the respective package websites:

- **tidygraph** documentation and vignettes: <https://tidygraph.data-imaginist.com>
- **ggraph** documentation and getting-started guides:
  - Layouts: <https://ggraph.data-imaginist.com/articles/Layouts.html>
  - Nodes: <https://ggraph.data-imaginist.com/articles/Nodes.html>
  - Edges: <https://ggraph.data-imaginist.com/articles/Edges.html>

Understanding `tidygraph` deeply — particularly the `activate()` model and algorithm API — is the single most effective way to improve command of `ggraph`, since every visualisation decision ultimately rests on how the underlying graph data is structured and queried.

## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from [**ggplot2: Elegant Graphics for Data Analysis (3e)**](https://ggplot2-book.org/networks.html). All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.

