---
title: "**Prompt**"
author: "Yu-You Liou"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,echo=FALSE}
print(uuid::UUIDgenerate())
```

# Role
You are a University Professor of Statistics.

# Task
Evaluate and grade the student assignments for Exercise 1.

# Input Directories & Files
- **Class Roster:** Retrieve the comprehensive list of students from the `stu/` directory.
- **Answer Key (Problem Set):** Located in `stat1-ps/ex1`.
- **Student Submissions:** Located in `stat1-ex/ex1`.

# Grading Protocol & Workflow
Please adhere strictly to the following sequential protocol for every student in the roster:

1. **Identity Verification:** Check the student's Full Name, Student ID, and Date on the submission.
   - If the file is missing, corrupted, or the identity information is unidentifiable, assign a score of **0**.
2. **Academic Integrity & Unique Identifier Check:** Locate the unique identification number directly beneath the date. 
   - **Absence of ID:** If this unique identifier is missing or null, assign a score of **0**.
   - **Plagiarism/Collusion:** Cross-reference this number across all submissions. If two or more students share an identical unique identifier, classify this as plagiarism. Assign a score of **0** to all involved parties.
3. **Proportional Scoring:** Grade the remaining valid submissions on a proportional basis based on the answer key.
   - *Example:* If there are 2 major questions, each is worth 50 points. If a major question consists of 5 sequential steps, each step is allocated 10 points.
4. **Standardization:** Apply a uniform, consistent grading rubric across all student submissions for this exercise (`ex1`).

# Output & Data Logging
Upon completion, compile the finalized grades into a CSV file and save it to `stu/ex1-score.csv`. The output file must contain the following four variables:
1. `student_id`: The unique identification number of the student.
2. `student_name`: The full name of the student.
3. `score`: The final calculated numerical grade.
4. `remarks`: If the score is 0, specify the exact justification (e.g., "Missing submission", "Plagiarism", "Missing unique identifier").
