---
title: "Prompt"
author: "Yu-You Liou"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,echo=FALSE}
print(uuid::UUIDgenerate())
```

## Role

You are an expert academic assistant and an experienced data science instructor proficient in R programming and the ggplot2 package.

## Task

Generate a comprehensive, plagiarism-free lecture slide for *Chapter 4* based on the official textbook **Elementary Statistics: A Step-by-Step Approach**.
(Available at `prompt/bluman-step-by-step-statistics-8th-edition`).

## Instructions

1.	**Structural Alignment**: Replicate the exact formatting, structural hierarchy, and stylistic layout of the provided *Chapter 3* slides (in `ch3`) to ensure visual and pedagogical consistency across chapters.

(*Note*: Do not include section numbers like x.x or x.x.x in slide titles, except for the first page.)

2.	**Exhaustive Coverage**: The handout must be meticulously detailed. Do not summarize or omit topics. Every single code chunk, function, argument, and conceptual point introduced in Chapter 4 of the textbook must be fully included and explained.

3.	**Paraphrasing & Academic Integrity**: To prevent plagiarism, you must completely rewrite and paraphrase all textual descriptions, explanations, and commentary from the textbook. The narrative flow and wording must be entirely original, academic, and tailored for a student audience, while strictly preserving the technical accuracy of the original content.

4.	**Code Execution**: Ensure all R code syntax is updated, correct, and properly formatted in markdown code blocks.

5. **Directory and Storage Architecture**: Replicate the established file structure utilized in the previous module. Dynamically generate and store the output within a newly designated directory (e.g., `ch3/`), mirroring the organization and naming conventions of the `ch3/` repository.
