---
title: "Data Analysis"
subtitle: "Chapter 4: Collective geoms"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "`ggplot2`: Elegant Graphics for Data Analysis"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch4-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

# Collective geoms

## Collective geoms

Geoms can be broadly classified as either **individual** or **collective**. An individual geom renders one distinct graphical object per row of data — for instance, `geom_point()` places a single point for each observation.

A **collective geom**, by contrast, represents multiple observations with a single geometric object. This behaviour may arise from a statistical summary (such as a boxplot) or may be intrinsic to how the geom is defined (such as a polygon).

Lines and paths occupy an intermediate position: each rendered line is built from a sequence of straight segments, yet each segment encodes two data points.

------------------------------------------------------------------------

The mechanism that governs how observations are assigned to graphical elements is the `group` aesthetic.

By default, `group` is derived from the interaction of all discrete variables mapped in the plot. This implicit grouping frequently produces the desired partitioning — but when it does not, or when no discrete variable is present, the grouping structure must be specified explicitly by mapping `group` to an appropriate variable.

------------------------------------------------------------------------

**The `Oxboys` dataset**

The examples throughout this chapter draw on the `Oxboys` dataset from the `nlme` package. It records the height (`height`) and centred age (`age`) of 26 boys (`Subject`), each measured on nine separate occasions (`Occasion`). Both `Subject` and `Occasion` are stored as ordered factors.

```{r,echo=T,warning=FALSE,out.width="80%",fig.align='center'}
library(ggplot2)
library(nlme)
data(Oxboys, package = "nlme")
head(Oxboys)
```

# Multiple groups, one aesthetic

## Multiple groups, one aesthetic

A common scenario is wanting to separate data into groups while rendering all groups with the same visual style. The goal is to make individual subjects distinguishable without labelling them. Longitudinal studies with many subjects frequently call for this treatment, and the resulting plots are informally known as **spaghetti plots**.

The plot below traces the growth trajectory of each boy over time. Setting `group = Subject` inside `aes()` instructs `geom_line()` to draw one connected line per subject:

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE,out.width="75%",fig.align='center'}
ggplot(Oxboys, aes(age, height, group = Subject)) +
  geom_point() +
  geom_line()
```

------------------------------------------------------------------------

**What happens without the correct grouping?**

If the `group` aesthetic is omitted, ggplot2 finds no discrete variable and treats all observations as belonging to a single group. The result is a characteristic **sawtooth pattern** — lines connecting observations from different subjects in an arbitrary order — which is clearly misleading:

```{r,echo=T,warning=FALSE,out.width="70%",fig.align='center'}
ggplot(Oxboys, aes(age, height)) +
  geom_point() +
  geom_line()
```

------------------------------------------------------------------------

**Grouping by multiple variables**

When a group is defined by the combination of more than one variable rather than a single column, use `interaction()` to construct a composite grouping key:

```{r,echo=T,eval=FALSE}
# Example: grouping by school and student together
aes(group = interaction(school_id, student_id))
```

This ensures that observations sharing the same school *and* the same student identifier are treated as belonging to the same group.

------------------------------------------------------------------------

**Exercises**

1. Draw a boxplot of `hwy` for each value of `cyl`, without converting `cyl` into a factor. What additional aesthetic must you set to make this work?

2. Modify the plot below so that a separate boxplot is produced for each integer value of `displ`:

```{r,echo=T,eval=FALSE}
ggplot(mpg, aes(displ, cty)) +
  geom_boxplot()
```

# Different groups on different layers

## Different groups on different layers

There are situations in which different layers of a plot require different levels of aggregation. For example, one layer may display individual trajectories while another overlays a single overall trend line.

If the grouping aesthetic is set globally in `ggplot()`, it propagates to every layer — including `geom_smooth()`. The result is one smoothed line per group, which is typically not what we want:

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
ggplot(Oxboys, aes(age, height, group = Subject)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE)
```

This produces one linear fit per subject, rather than a single trend line summarising all boys together. Remember: grouping controls not only the geometry but also the statistical transformations — each stat runs independently for each group.

------------------------------------------------------------------------

**Fixing the problem: layer-specific grouping**

The solution is to move the `group` aesthetic from the global `ggplot()` call into `geom_line()` alone. The `geom_smooth()` layer then has no discrete variable to group by, so it defaults to treating all observations as one group and produces a single overall smooth:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="65%",fig.align='center'}
ggplot(Oxboys, aes(age, height)) +
  geom_line(aes(group = Subject)) +
  geom_smooth(method = "lm", linewidth = 2, se = FALSE)
```

------------------------------------------------------------------------

**Key takeaway**

Aesthetics set inside `ggplot()` are **global** and apply to every subsequent layer. Aesthetics set inside an individual geom function are **local** and apply only to that layer. This distinction is critical when different layers require different grouping structures.

# Overriding the default grouping

## Overriding the default grouping

Sometimes a plot has a discrete x-axis, yet we still want to draw lines that connect observations *across* groups rather than within them. This pattern appears in interaction plots, profile plots, and parallel coordinate plots, among others.

Consider boxplots of `height` at each measurement `Occasion`. Since `Occasion` is a discrete variable, ggplot2 automatically draws one boxplot per occasion:

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
ggplot(Oxboys, aes(Occasion, height)) +
  geom_boxplot()
```

------------------------------------------------------------------------

**Naïvely adding lines does not work**

Simply adding `geom_line()` to the boxplot produces lines drawn *within* each occasion (connecting observations at the same x position), not *across* occasions per subject. This is because the default grouping is driven by `Occasion`:

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
ggplot(Oxboys, aes(Occasion, height)) +
  geom_boxplot() +
  geom_line(colour = "#3366FF", alpha = 0.5)
```

------------------------------------------------------------------------

**Overriding the grouping in a single layer**

The fix is to explicitly override the grouping in `geom_line()` to specify one line per subject. The boxplot layer retains its default grouping by `Occasion`, while the line layer uses `Subject`:

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
ggplot(Oxboys, aes(Occasion, height)) +
  geom_boxplot() +
  geom_line(aes(group = Subject), colour = "#3366FF", alpha = 0.5)
```

Each blue line now traces an individual boy's growth across all nine occasions, exactly as intended.

------------------------------------------------------------------------

**Exercises**

3. When illustrating the difference between mapping a continuous vs. a discrete colour to a line, the discrete example required `aes(group = 1)`. Why? What happens when it is omitted? What is the difference between `aes(group = 1)` and `aes(group = 2)`?

# Matching aesthetics to graphic objects

## Matching aesthetics to graphic objects

An important question arises with collective geoms: when individual observations carry different aesthetic values, how should those values be mapped onto a single geometric object?

ggplot2 handles this differently depending on the geom type.

------------------------------------------------------------------------

**Lines and paths: the "first value" rule**

For `geom_line()` and `geom_path()`, each segment is defined by two consecutive observations. ggplot2 applies the aesthetic value (e.g., `colour`) from the **first** of the two observations to each segment. The aesthetic of the final observation is never applied to any segment.

```{r,echo=T,warning=FALSE,out.width="70%",fig.align='center'}
library(patchwork)
df <- data.frame(x = 1:3, y = 1:3, colour = c(1, 3, 5))

p1 <- ggplot(df, aes(x, y, colour = factor(colour))) +
  geom_line(aes(group = 1), linewidth = 2) +
  geom_point(size = 5)

p2 <- ggplot(df, aes(x, y, colour = colour)) +
  geom_line(aes(group = 1), linewidth = 2) +
  geom_point(size = 5)

p1 | p2
```

------------------------------------------------------------------------

In the left panel (discrete colour), the first point and first segment are red, the second point and second segment are green, and the last point (no segment follows it) is blue.

In the right panel (continuous colour), the same "first value" principle applies to three shades of blue. Crucially, ggplot2 does **not** smoothly interpolate between aesthetic values — even for a continuous variable. If a smooth colour gradient along the line is desired, the interpolation must be performed manually:

------------------------------------------------------------------------

**Manual linear interpolation for smooth colour gradients**

```{r,echo=T,warning=FALSE,out.width="70%",fig.align='center'}
xgrid <- with(df, seq(min(x), max(x), length = 50))
interp <- data.frame(
  x      = xgrid,
  y      = approx(df$x, df$y,      xout = xgrid)$y,
  colour = approx(df$x, df$colour, xout = xgrid)$y
)
ggplot(interp, aes(x, y, colour = colour)) +
  geom_line(linewidth = 2) +
  geom_point(data = df, size = 5)
```

The `approx()` function linearly interpolates both the `y` coordinate and the `colour` value at 50 evenly spaced positions, producing a smooth gradient along the path.

------------------------------------------------------------------------

**Additional constraint: constant line type**

An important limitation applies to paths and lines: the `linetype` aesthetic must remain constant over the entire line. R has no mechanism for rendering a line whose type (solid, dashed, dotted, etc.) varies continuously along its length.

------------------------------------------------------------------------

**Polygons and other complex collective geoms**

For geoms more complex than lines — such as polygons — a single geometric object may map onto many observations, making it ambiguous how differing aesthetic values should be reconciled. For example, how would a polygon be filled if each of its border vertices specifies a different fill colour?

ggplot2 resolves this ambiguity with a simple rule: **aesthetics from the individual observations are used only when they are all identical across the group**. If the aesthetic values differ, ggplot2 falls back to its default value instead.

------------------------------------------------------------------------

**Discrete variables: splitting into sub-groups**

For **discrete** fill or colour variables, ggplot2 incorporates the variable into the group aesthetic automatically. This splits the collective geom into smaller pieces. For bar and area charts, stacking these pieces preserves the shape of the ungrouped data:

```{r,echo=T,warning=FALSE,out.width="70%",fig.align='center'}
p1 <- ggplot(mpg, aes(class)) +
  geom_bar()
p2 <- ggplot(mpg, aes(class, fill = drv)) +
  geom_bar()
p1 | p2
```

------------------------------------------------------------------------

**Continuous fill with bars: a common pitfall**

Mapping a **continuous** variable to `fill` in `geom_bar()` does not work the same way. Because each bar spans multiple observations with different `hwy` values, no single colour can summarise them, and ggplot2 reverts to the default grey:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="65%",fig.align='center'}
ggplot(mpg, aes(class, fill = hwy)) +
  geom_bar()
```

------------------------------------------------------------------------

**Fixing it: override the grouping**

To display multiple fill colours within each bar, we must override the grouping so that each unique `hwy` value forms its own sub-bar. The stacked result reveals the distribution of highway fuel economy within each vehicle class:

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
ggplot(mpg, aes(class, fill = hwy, group = hwy)) +
  geom_bar()
```

When this technique is used, the bars are stacked in the order defined by the grouping variable (here, `hwy`). If precise control over the stacking order is needed, convert `hwy` to a factor with levels arranged accordingly.

------------------------------------------------------------------------

**Exercises**

4. How many bars are displayed in each of the following plots? Explain your reasoning.

```{r,echo=T,eval=FALSE}
ggplot(mpg, aes(drv)) +
  geom_bar()

ggplot(mpg, aes(drv, fill = hwy, group = hwy)) +
  geom_bar()

library(dplyr)
mpg2 <- mpg |> arrange(hwy) |> mutate(id = seq_along(hwy))
ggplot(mpg2, aes(drv, fill = hwy, group = id)) +
  geom_bar()
```

*(Hint: add `colour = "white"` to outline individual sub-bars.)*

5. Install the `babynames` package and run the code below. Identify what is wrong with the resulting graph and fix it. Why does the uncorrected version look unexpected?

```{r,echo=T,eval=FALSE}
library(babynames)
hadley <- dplyr::filter(babynames, name == "Hadley")
ggplot(hadley, aes(year, n)) +
  geom_line()
```

## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from [**ggplot2: Elegant Graphics for Data Analysis (3e)**](https://ggplot2-book.org/collective-geoms.html). All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.

