---
title: "Data Analysis"
subtitle: "Chapter 13: Build a Plot Layer by Layer"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "`ggplot2`: Elegant Graphics for Data Analysis"
logo: "logo.svg"
editor: source
bibliography: bib.bib
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch13-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

# Build a Plot Layer by Layer

## Overview

One of the key ideas behind `ggplot2` is that it lets you **iterate**, building up a complex plot a layer at a time. Each layer can draw on a different dataset and a different aesthetic mapping, making it possible to combine data from multiple sources in a single graphic.

You have already created layers with functions like `geom_point()` and `geom_histogram()`. This chapter dissects a layer and shows how to control all **five** of its components:

- **Data** — the data frame the layer draws from
- **Aesthetic mappings** — how variables connect to visual properties
- **Geom** — the geometric object that renders each observation
- **Stat** — the statistical transformation applied before drawing
- **Position adjustment** — how overlapping elements are nudged apart

# Building a Plot

## Two Distinct Steps

Whenever we create a plot with `ggplot()` and immediately add a geom, two distinct steps are actually happening. First, we create a plot with a default dataset and aesthetic mappings:

```{r, echo=TRUE, warning=FALSE, message=FALSE, out.width="50%", fig.align='center'}
library(ggplot2)

p <- ggplot(mpg, aes(displ, hwy))
p
```

There is nothing to see yet — we have specified the data and mappings but no geometric object to draw them.

---

We then add a layer to make something appear:

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
p + geom_point()
```

`geom_point()` is a **shortcut**. Behind the scenes it calls the `layer()` function to build a new layer with all five components fully specified.

---

The explicit `layer()` call makes the five components visible:

```{r, echo=TRUE, eval=FALSE}
p + layer(
  mapping  = NULL,        # inherit aes() from ggplot()
  data     = NULL,        # inherit data from ggplot()
  geom     = "point",     # geometric object
  stat     = "identity",  # statistical transformation
  position = "identity"   # position adjustment
)
```

You will rarely write `layer()` in full because it is so verbose. Instead, the `geom_` functions are convenient shortcuts: `geom_point(mapping, data, ...)` is exactly equivalent to `layer(mapping, data, geom = "point", ...)`. Understanding `layer()` nonetheless gives you a clearer mental model of what a layer really is.

# Data

## Data

Every layer must have data associated with it, and that data must be a **tidy data frame** — variables in columns, observations in rows. This is a strong restriction, but it has good reasons: your data is explicit, a single data frame is easy to save and share, and it enforces a clean separation between *making* data and *visualising* it.

Crucially, the data on each layer **need not be the same**. Combining several datasets in one plot is often useful. To illustrate, we generate two datasets derived from `mpg`: a loess fit and a set of outliers.

---

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(tibble)
library(dplyr)

mod  <- loess(hwy ~ displ, data = mpg)
grid <- tibble(displ = seq(min(mpg$displ), max(mpg$displ), length = 50))
grid$hwy <- predict(mod, newdata = grid)

std_resid <- resid(mod) / mod$s
outlier   <- filter(mpg, abs(std_resid) > 2)
outlier[, c("model", "displ", "hwy")]
```

---

With these new datasets we can enhance the raw scatterplot — overlaying a smoothed line and labelling the outlying points. Note that each layer supplies its own `data =`:

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_line(data = grid, colour = "blue", linewidth = 1.5) +
  geom_text(data = outlier, aes(label = model))
```

---

You need the explicit `data =` in the layers, but not in the call to `ggplot()`. This is because the argument order differs: `ggplot(data, mapping)` versus `geom_*(mapping, data)`. The inconsistency reduces typing for the common case of specifying the data once in `ggplot()`.

The same plot can be written without a default dataset, supplying data to every layer instead:

```{r, echo=TRUE, eval=FALSE}
ggplot(mapping = aes(displ, hwy)) +
  geom_point(data = mpg) +
  geom_line(data = grid) +
  geom_text(data = outlier, aes(label = model))
```

This style is less clear when there is a primary dataset, but you may prefer it when no single dataset dominates.

# Aesthetic Mappings

## Aesthetic Mappings

The aesthetic mappings, defined with `aes()`, describe how variables map to visual properties:

```{r, echo=TRUE, eval=FALSE}
aes(x = displ, y = hwy, colour = class)
```

The first two arguments default to `x` and `y`, so `aes(displ, hwy, colour = class)` is equivalent. A few guidelines:

- Keep calculations in `aes()` simple; move complex transformations into an explicit `dplyr::mutate()`.
- **Never** refer to a variable with `$` (e.g. `diamonds$carat`) inside `aes()`. This breaks containment and fails when ggplot2 reorders rows (as it does when faceting).

---

### Plot vs. Layer Aesthetics

Mappings can be supplied in `ggplot()`, in individual layers, or both. All of these produce the same specification:

```{r, echo=TRUE, eval=FALSE}
ggplot(mpg, aes(displ, hwy, colour = class)) + geom_point()
ggplot(mpg, aes(displ, hwy)) + geom_point(aes(colour = class))
ggplot(mpg, aes(displ)) + geom_point(aes(y = hwy, colour = class))
ggplot(mpg) + geom_point(aes(displ, hwy, colour = class))
```

Within a layer you can **add**, **override**, or **remove** mappings:

| Operation | Layer aesthetics | Result |
|---|---|---|
| Add | `aes(colour = cyl)` | `aes(displ, hwy, colour = cyl)` |
| Override | `aes(y = cty)` | `aes(displ, cty)` |
| Remove | `aes(y = NULL)` | `aes(displ)` |

---

The distinction matters once you add multiple layers. These two plots are both valid but emphasise different things — the left colours *both* the points and the smooth by class; the right colours only the points:

```{r, echo=TRUE, warning=FALSE, message=FALSE, out.width="50%", fig.align='center'}
library(patchwork)
p1 <- ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE) +
  theme(legend.position = "none")
p2 <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) + geom_smooth(method = "lm", se = FALSE) +
  theme(legend.position = "none")
p1 | p2
```

---

### Setting vs. Mapping

Instead of **mapping** an aesthetic to a variable, you can **set** it to a single constant value as a layer parameter. The rule:

- To govern appearance by a variable → put it **inside** `aes()`.
- To override a fixed colour or size → put the value **outside** `aes()`.

```{r, echo=TRUE, warning=FALSE, out.width="48%", fig.align='center'}
s1 <- ggplot(mpg, aes(cty, hwy)) + geom_point(colour = "darkblue")
s2 <- ggplot(mpg, aes(cty, hwy)) + geom_point(aes(colour = "darkblue"))
s1 | s2
```

---

The second plot above **maps** colour to the literal string `"darkblue"`. This creates a one-level variable that the default discrete scale renders as a pinkish hue — almost certainly not what you wanted.

Mapping to constants is, however, genuinely useful for **naming** layers so they appear in a legend:

```{r, echo=TRUE, warning=FALSE, message=FALSE, out.width="55%", fig.align='center'}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(aes(colour = "loess"), method = "loess", se = FALSE) +
  geom_smooth(aes(colour = "lm"), method = "lm", se = FALSE) +
  labs(colour = "Method")
```

# Geoms

## Geoms

Geometric objects, or **geoms**, perform the actual rendering of a layer and determine the type of plot. A point geom makes a scatterplot; a line geom makes a line plot. ggplot2 ships with a large vocabulary of geoms, organised by the number of variables they display:

- **Graphical primitives**: `geom_blank()`, `geom_point()`, `geom_path()`, `geom_ribbon()`, `geom_segment()`, `geom_rect()`, `geom_polygon()`, `geom_text()`
- **One variable**: `geom_bar()`, `geom_histogram()`, `geom_density()`, `geom_dotplot()`, `geom_freqpoly()`
- **Two variables**: `geom_point()`, `geom_smooth()`, `geom_boxplot()`, `geom_bin2d()`, `geom_hex()`, `geom_count()`, `geom_jitter()`, `geom_area()`, `geom_line()`, `geom_step()`
- **Three variables**: `geom_contour()`, `geom_tile()`, `geom_raster()`

---

Each geom understands a set of aesthetics, some of which are **required**. A point requires `x` and `y` and understands `colour`, `size`, and `shape`; a bar requires a height (`ymax`) and understands `width`, border colour, and fill.

Some geoms differ mainly in how they are **parameterised**. You can draw the same square three ways:

| Geom | Parameterisation |
|---|---|
| `geom_tile()` | location (`x`, `y`) + dimensions (`width`, `height`) |
| `geom_rect()` | corners (`xmin`, `xmax`, `ymin`, `ymax`) |
| `geom_polygon()` | a four-row data frame of corner `x`/`y` positions |

Picking the parameterisation that matches your data usually makes a plot much easier to draw.

# Stats

## Stats

A statistical transformation, or **stat**, transforms the data — typically by summarising it. Many familiar geoms are powered by a stat behind the scenes:

| Stat | Geoms |
|---|---|
| `stat_bin()` | `geom_bar()`, `geom_freqpoly()`, `geom_histogram()` |
| `stat_bin2d()` | `geom_bin2d()` |
| `stat_boxplot()` | `geom_boxplot()` |
| `stat_smooth()` | `geom_smooth()` |
| `stat_sum()` | `geom_count()` |

Other stats have no `geom_` shortcut, e.g. `stat_ecdf()`, `stat_function()`, `stat_summary()`, `stat_qq()`, `stat_unique()`.

---

You can either add a `stat_()` and override its geom, or add a `geom_()` and override its stat. Both produce the same red mean marker:

```{r, echo=TRUE, warning=FALSE, out.width="48%", fig.align='center'}
m1 <- ggplot(mpg, aes(trans, cty)) +
  geom_point() +
  stat_summary(geom = "point", fun = "mean", colour = "red", size = 4)
m2 <- ggplot(mpg, aes(trans, cty)) +
  geom_point() +
  geom_point(stat = "summary", fun = "mean", colour = "red", size = 4)
m1 | m2
```

The second form is preferable: it makes clear that you are displaying a **summary**, not the raw data.

---

### Generated Variables

A stat takes a data frame and returns a data frame, often adding **new variables** you can map to aesthetics. For example, `stat_bin()` (used by histograms) generates `count`, `density`, and `x`. To map a generated variable, wrap its name in `after_stat()`:

```{r, echo=TRUE, warning=FALSE, out.width="48%", fig.align='center'}
h1 <- ggplot(diamonds, aes(price)) + geom_histogram(binwidth = 500)
h2 <- ggplot(diamonds, aes(price)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 500)
h1 | h2
```

---

`after_stat()` prevents confusion when the raw data already contains a variable of the same name, and signals to any reader that the variable was stat-generated.

Standardising to density is especially useful for comparing groups of very different sizes. Compare counts versus densities of `price` within `cut`:

```{r, echo=TRUE, warning=FALSE, out.width="48%", fig.align='center'}
f1 <- ggplot(diamonds, aes(price, colour = cut)) +
  geom_freqpoly(binwidth = 500) + theme(legend.position = "none")
f2 <- ggplot(diamonds, aes(price, colour = cut)) +
  geom_freqpoly(aes(y = after_stat(density)), binwidth = 500) +
  theme(legend.position = "none")
f1 | f2
```

The density view reveals a surprise: low-quality diamonds appear *more* expensive on average.

# Position Adjustments

## Position Adjustments

Position adjustments apply minor tweaks to the positions of elements within a layer. Three apply mainly to **bars**:

- `position_stack()` — stack overlapping bars on top of each other (the default for bars)
- `position_fill()` — stack and rescale so each bar reaches 1
- `position_dodge()` — place overlapping bars side by side

```{r, echo=TRUE, warning=FALSE, out.width="32%", fig.align='center'}
a <- dplot <- ggplot(diamonds, aes(color, fill = cut)) +
  xlab(NULL) + ylab(NULL) + theme(legend.position = "none")
b <- dplot + geom_bar()
c <- dplot + geom_bar(position = "fill")
d <- dplot + geom_bar(position = "dodge")
```

---

```{r, echo=TRUE, warning=FALSE, out.width="32%", fig.align='center'}
(a|b)/(c|d)
```

---

There is also `position_identity()`, which does nothing. It is useless for bars (each obscures those behind it) but fine for geoms like lines that do not need adjusting.

Three position adjustments apply mainly to **points**:

- `position_nudge()` — move points by a fixed offset
- `position_jitter()` — add random noise to every position
- `position_jitterdodge()` — dodge within groups, then add noise

---

Position adjustments take parameters differently from stats and geoms: you construct a position object and supply arguments to it.

```{r, echo=TRUE, warning=FALSE, out.width="48%", fig.align='center'}
j1 <- ggplot(mpg, aes(displ, hwy)) + geom_point(position = "jitter")
j2 <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(position = position_jitter(width = 0.05, height = 0.5))
j1 | j2
```

The verbose form has a shortcut: `geom_jitter(width = 0.05, height = 0.5)`. Jittering is most useful for **discrete** data, where exact overlaps are common.

---

**Exercises**

1. The first two arguments to `ggplot()` are `data` then `mapping`; for every layer function they are `mapping` then `data`. Why does the order differ? What does this tell you about what you set most commonly?

2. Using `geom_line(data = grid)` and `geom_text(data = outlier)`, reproduce the enhanced `mpg` scatterplot from §13.3. Then rewrite it specifying data in every layer instead of in `ggplot()`.

3. Explain the difference in output between `geom_point(colour = "darkblue")` and `geom_point(aes(colour = "darkblue"))`. How does adding `scale_colour_identity()` change the second?

---

4. The default histogram maps bar height to `count`. Rewrite it to use `after_stat(density)` and explain when each is appropriate.

5. Recreate a red "mean" marker on top of a scatterplot of `cty` by `trans`, once with `stat_summary(geom = "point", ...)` and once with `geom_point(stat = "summary", ...)`. Which form communicates intent more clearly?

6. Build a stacked, a filled, and a dodged bar chart of `cut` within `color` from `diamonds`. When is each position adjustment most useful?

7. Why might you use `geom_jitter()` instead of `geom_count()` for overplotted discrete data? What are the trade-offs?

## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from [**ggplot2: Elegant Graphics for Data Analysis (3e)**](https://ggplot2-book.org/layers.html). All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.

## References
