---
title: "Data Analysis"
subtitle: "Chapter 10: Position Scales and Axes"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "`ggplot2`: Elegant Graphics for Data Analysis"
logo: "logo.svg"
editor: source
bibliography: bib.bib
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch10-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

# Position Scales and Axes

## Overview

Every plot produced by `ggplot2` relies on **position scales** to map data values to physical locations on the plotting surface. The x and y aesthetics each have an associated position scale, even when the user does not explicitly declare them.

Consider this minimal histogram specification:

```{r, echo=TRUE, eval=FALSE}
ggplot(mpg, aes(x = displ)) + geom_histogram()
```

The y aesthetic is never mentioned, yet the plot renders correctly because `geom_histogram()` internally computes a `count` variable and maps it to y. The fully explicit equivalent is:

```{r, echo=TRUE, eval=FALSE}
ggplot(mpg, aes(x = displ, y = after_stat(count))) + geom_histogram()
```

---

Because position scales appear in **every** plot, a thorough understanding of how they work—and how they can be customised—is essential for effective data visualisation.

This chapter is organised into four major sections:

- **§10.1** — Numeric (continuous) position scales: limits, zooming, expansion, breaks, minor breaks, labels, and transformations
- **§10.2** — Date-time position scales: specialised breaks and label formatting
- **§10.3** — Discrete position scales: reordering, custom labels, and axis guides
- **§10.4** — Binned position scales: slicing continuous variables into discrete bins

# Numeric Position Scales

## Numeric Position Scales

The default continuous position scales, `scale_x_continuous()` and `scale_y_continuous()`, establish a **linear mapping** from raw data values to pixel coordinates on the plot panel.

A family of convenience wrappers—`scale_x_log10()`, `scale_x_reverse()`, `scale_x_sqrt()`, and their y-axis counterparts—provides quick access to the most commonly needed transformations (covered fully in §10.1.7).

The fundamental behaviour of all continuous position scales is governed by six key arguments:

---

| Argument | Purpose |
|---|---|
| `limits` | Define the range of the scale |
| `expand` | Pad the visual range beyond the data |
| `breaks` | Set major tick positions |
| `minor_breaks` | Set minor tick positions |
| `labels` | Control tick label text |
| `trans` | Apply a mathematical transformation |

---

### Limits

The `limits` argument controls the range over which a continuous scale is defined. By default, ggplot2 derives limits automatically from the data range, but manual specification is often necessary when comparing across multiple plots.

```{r, echo=TRUE, warning=FALSE, message=FALSE, out.width="70%", fig.align='center'}
library(ggplot2)
library(dplyr)

# Two subsets of mpg that will have different natural axis ranges
mpg_99 <- mpg |> filter(year == 1999)
mpg_08 <- mpg |> filter(year == 2008)

base_99 <- ggplot(mpg_99, aes(displ, hwy)) + geom_point()
base_08 <- ggplot(mpg_08, aes(displ, hwy)) + geom_point()
```

Without explicit limits, each plot uses its own natural data range, making side-by-side comparison unreliable.

---

To synchronise the axes, pass a length-2 numeric vector to `limits` in each scale. Use `NA` as a placeholder if you only want to constrain one endpoint:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
library(patchwork)
p99 <- base_99 +
  scale_x_continuous(limits = c(1, 7)) +
  scale_y_continuous(limits = c(10, 45))
p08 <- base_08 +
  scale_x_continuous(limits = c(1, 7)) +
  scale_y_continuous(limits = c(10, 45))
p99 | p08
```

---

The `lims()` convenience function reduces verbosity when setting limits on multiple aesthetics simultaneously:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
p99b <- base_99 + lims(x = c(1, 7), y = c(10, 45))
p08b <- base_08 + lims(x = c(1, 7), y = c(10, 45))
p99b | p08b
```

`lims()` follows the same name-value syntax as `labs()`: the argument name specifies the aesthetic, and the value specifies its limits. Single-axis helpers `xlim()` and `ylim()` are also available for quick adjustments.

---

### Zooming In

Narrowing the scale `limits` is **not** the same as zooming in visually. When a data value falls outside the declared limits of a continuous scale, ggplot2 converts it to `NA` **before** statistical summaries are computed. This can silently distort the output of stats such as `stat_boxplot`.

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
base <- ggplot(mpg, aes(drv, hwy)) +
  geom_hline(yintercept = 28, colour = "red") +
  geom_boxplot()
base
```

---

Comparing the two approaches to restricting the visible range:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
# Correct: zoom without dropping data
p_zoom <- base + coord_cartesian(ylim = c(10, 35))

# Incorrect: drops data before stat computation
p_lim  <- base + ylim(10, 35)   # Warning: rows removed

p_zoom | p_lim
```

The left panel zooms safely. The right panel has silently recomputed the boxplot on a truncated sample, shifting the median downward. **Always use `coord_cartesian()` when the intent is purely visual zoom.**

---

### Visual Range Expansion

By default, ggplot2 adds a small buffer beyond the data limits so that plotted objects do not collide with the axis lines. This buffer is controlled via the `expand` argument, which accepts an `expansion()` call.

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
base_raster <- ggplot(faithfuld, aes(waiting, eruptions)) +
  geom_raster(aes(fill = density)) +
  theme(legend.position = "none") +
  labs(x = NULL, y = NULL)

# Remove all padding (useful for raster/image plots)
base_raster +
  scale_x_continuous(expand = expansion(0)) +
  scale_y_continuous(expand = expansion(0))
```

---

`expansion()` accepts two arguments:

- `add` — a constant amount of space in data units (same scale as the variable)
- `mult` — a proportional amount relative to the full axis range

Both can be length-1 (symmetric) or length-2 (asymmetric: lower, upper).

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
base_raster +
  scale_x_continuous(expand = expansion(add = 3)) +
  scale_y_continuous(expand = expansion(add = 3))
```

---

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
# 20% multiplicative padding on all sides
base_raster +
  scale_x_continuous(expand = expansion(mult = .2)) +
  scale_y_continuous(expand = expansion(mult = .2))
```

---

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
# Asymmetric: 5% at the low end, 20% at the high end
base_raster +
  scale_x_continuous(expand = expansion(mult = c(.05, .2))) +
  scale_y_continuous(expand = c(.05, 0, .2, 0))
```

---

### Breaks

Axis tick marks in ggplot2 are a special case of **scale breaks**. Their positions are set via the `breaks` argument to the relevant scale function.

---

```{r, echo=TRUE, warning=FALSE, out.width="70%", fig.align='center'}
# Toy dataset used throughout this section
toy <- data.frame(
  const = 1,
  up    = 1:4,
  txt   = letters[1:4],
  big   = (1:4) * 1000,
  log   = c(2, 5, 10, 2000)
)

base <- ggplot(toy, aes(big, const)) +
  geom_point() +
  labs(x = NULL, y = NULL) +
  scale_y_continuous(breaks = NULL)   # suppresses y-axis ticks
base
```

Setting `breaks = NULL` removes tick marks **and** the associated major grid lines.

---

Manually specifying a vector of break positions moves both the major grid lines and the minor grid lines that sit between them:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
p1 <- base + scale_x_continuous(breaks = c(1000, 2000, 4000))
p2 <- base + scale_x_continuous(breaks = c(1000, 1500, 2000, 4000))
p1 / p2
```

---

The `breaks` argument also accepts a **function** that receives the scale limits and returns break positions. The `scales` package provides several ready-made break functions:

| Function | Description |
|---|---|
| `scales::breaks_extended()` | Automatic "nice" breaks (ggplot2 default) |
| `scales::breaks_log()` | Breaks suited for log-transformed axes |
| `scales::breaks_pretty()` | Aesthetically pleasing breaks for date-times |
| `scales::breaks_width()` | Equally-spaced breaks at a fixed width |

---

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
p_def <- base
p_ext <- base + scale_x_continuous(breaks = scales::breaks_extended())
p_few <- base + scale_x_continuous(breaks = scales::breaks_extended(n = 2))
p_def | p_ext | p_few
```

---

`breaks_width()` places breaks at a fixed interval. An optional `offset` shifts all breaks by a constant amount:

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
p_w1 <- base + scale_x_continuous(breaks = scales::breaks_width(800))
p_w2 <- base + scale_x_continuous(breaks = scales::breaks_width(800, offset =  200))
p_w3 <- base + scale_x_continuous(breaks = scales::breaks_width(800, offset = -200))
p_w1 | p_w2 | p_w3
```

Note how a positive offset shifts all break positions upward, while a negative offset shifts them downward.

---

### Minor Breaks

Minor breaks are the faint, unlabelled grid lines that appear between the labelled major breaks. They are set via the `minor_breaks` argument, which also accepts a numeric vector or a function.

Minor breaks are especially informative on **log-transformed axes**, where they provide a clear visual cue that the spacing is non-linear:

---

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
# Build a dense set of minor break positions across several decades
mb <- unique(as.numeric(1:10 %o% 10^(0:3)))

base_log <- ggplot(toy, aes(log, const)) +
  geom_point() +
  labs(x = NULL, y = NULL) +
  scale_y_continuous(breaks = NULL)

p_no_mb <- base_log + scale_x_log10()
p_mb    <- base_log + scale_x_log10(minor_breaks = mb)
p_no_mb | p_mb
```

The `scales` package also supplies `scales::minor_breaks_n()` and `scales::minor_breaks_width()` as function-based alternatives.

---

### Labels

Every break has an associated text label. Labels are controlled by the `labels` argument, which accepts either a character vector (one label per break) or a labelling function.

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
base_lbl <- ggplot(toy, aes(big, const)) +
  geom_point() +
  labs(x = NULL, y = NULL) +
  scale_y_continuous(breaks = NULL)

p_default <- base_lbl
p_custom  <- base_lbl +
  scale_x_continuous(breaks = c(2000, 4000), labels = c("2k", "4k"))
p_default | p_custom
```

---

The `scales` package offers a suite of ready-made label formatters:

| Function | Output example |
|---|---|
| `scales::label_bytes()` | 1.5 kB, 2.3 MB |
| `scales::label_comma()` | 1,000 |
| `scales::label_dollar()` | $1,000 |
| `scales::label_ordinal()` | 1st, 2nd, 3rd |
| `scales::label_percent()` | 10% |
| `scales::label_pvalue()` | <.05, .34 |

---

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
base_lbl2 <- ggplot(toy, aes(big, const)) +
  geom_point() +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(breaks = NULL)

p_pct <- base_lbl2 + scale_y_continuous(labels = scales::label_percent())
p_eur <- base_lbl2 + scale_y_continuous(
  labels = scales::label_dollar(prefix = "", suffix = "€")
)
p_pct | p_eur
```

---

There is an important difference between setting `breaks = NULL` and `labels = NULL`:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
p_brk_null <- base_lbl2 + scale_y_continuous(breaks = NULL)
p_lbl_null <- base_lbl2 + scale_y_continuous(labels = NULL)
p_brk_null | p_lbl_null
```

- `breaks = NULL` removes both the tick marks **and** the major grid lines.
- `labels = NULL` hides only the text labels while preserving ticks and grid lines.

---

### Transformations

By default, continuous position scales use an **identity (linear)** transformation. Overriding this is done via the `trans` argument, which accepts either a transformation name (string) or a transformer object from the `scales` package.

Convenience wrappers such as `scale_x_log10()`, `scale_x_reverse()`, and `scale_x_sqrt()` cover the most common cases:

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
base_trans <- ggplot(mpg, aes(displ, hwy)) + geom_point()
p_rev_x <- base_trans + scale_x_reverse()
p_rev_y <- base_trans + scale_y_reverse()
base_trans | p_rev_x | p_rev_y
```

---

For less common transformations, use the `trans` argument directly:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
p_recip <- ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(trans = "reciprocal")

p_log <- ggplot(diamonds, aes(price, carat)) +
  geom_bin2d() +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10")

p_recip | p_log
```

---

The full list of built-in transformations supplied by the `scales` package:

| Name | Function $f(x)$ | Inverse $f^{-1}(x)$ |
|---|---|---|
| `"asn"` | $\tanh^{-1}(x)$ | $\tanh(y)$ |
| `"exp"` | $e^x$ | $\log(y)$ |
| `"identity"` | $x$ | $y$ |
| `"log"` | $\log(x)$ | $e^y$ |
| `"log10"` | $\log_{10}(x)$ | $10^y$ |
| `"log2"` | $\log_2(x)$ | $2^y$ |
| `"logit"` | $\log\!\left(\frac{x}{1-x}\right)$ | $\frac{1}{1+e^y}$ |
| `"probit"` | $\Phi(x)$ | $\Phi^{-1}(y)$ |
| `"reciprocal"` | $x^{-1}$ | $y^{-1}$ |
| `"reverse"` | $-x$ | $-y$ |
| `"sqrt"` | $x^{1/2}$ | $y^2$ |

---

You can also call the transformer object directly—these two specifications are equivalent:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
# Using the name string
a <- ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(trans = "reciprocal")

# Using the transformer object
b <- ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(trans = scales::reciprocal_trans())

a | b
```

---

Similarly, the convenience wrapper and the explicit `trans` argument are interchangeable:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
# Via convenience wrapper
a <- ggplot(diamonds, aes(price, carat)) +
  geom_bin2d() +
  scale_x_log10() +
  scale_y_log10()

# Via the trans argument
b <- ggplot(diamonds, aes(price, carat)) +
  geom_bin2d() +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10")

a | b
```

---

**Transforming the scale vs. transforming the data**

An important subtlety: transforming the scale is different from transforming the data variable directly.

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
# Axes labelled on the *transformed* (log10) scale
p_data <- ggplot(mpg, aes(log10(displ), hwy)) + geom_point() +
  ggtitle("log10(displ) in aes()")

# Axes labelled on the *original* data scale
p_scale <- ggplot(mpg, aes(displ, hwy)) + geom_point() +
  scale_x_log10() +
  ggtitle("scale_x_log10()")

p_data | p_scale
```

---

In both cases the statistical geometry is identical, but the axis tick labels differ. Transforming the **scale** (right) displays original data values on the ticks, which is generally more interpretable. Importantly, in both approaches the transformation occurs *before* any statistical summaries—use `coord_trans()` if transformation is needed *after* statistics.

# Date-Time Position Scales

## Date-Time Position Scales

When a date or date-time variable is mapped to x or y, ggplot2 automatically selects `scale_x_date()` or `scale_x_datetime()` as the default scale. These scales extend the continuous scale framework with date-friendly arguments for controlling breaks and labels.

Supported R date/time classes include:

- `Date` — calendar dates (base R)
- `POSIXct` — date-time values (base R)
- `hms` — time-of-day values (hms package)

Use `as.Date()`, `as.POSIXct()`, or `hms::as_hms()` to convert character strings to the appropriate class before plotting.

---

### Breaks for Date Scales

The `date_breaks` argument positions tick marks using human-readable date units: `"years"`, `"months"`, `"weeks"`, `"days"`, `"hours"`, `"minutes"`, or `"seconds"`. A numeric prefix sets the interval size.

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
date_base <- ggplot(economics, aes(date, psavert)) +
  geom_line(na.rm = TRUE) +
  labs(x = NULL, y = NULL)

p_auto <- date_base
p_15yr <- date_base + scale_x_date(date_breaks = "15 years")
p_auto / p_15yr
```

---

`date_breaks = "15 years"` is shorthand for `breaks = scales::breaks_width("15 years")`. Using the longer form grants access to the `offset` argument:

```{r, echo=TRUE, eval=FALSE}
# Breaks on the 9th of each month (salary cycle example)
the_year  <- as.Date(c("2021-01-01", "2021-12-31"))
set_breaks <- scales::breaks_width("1 month", offset = 8)
set_breaks(the_year)
#>  [1] "2021-01-09" "2021-02-09" ... "2021-12-09" "2022-01-09"
```

---

### Minor Breaks for Date Scales

The `date_minor_breaks` argument controls minor break spacing in exactly the same way as `date_breaks` controls major break spacing.

---

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
df <- data.frame(y = as.Date(c("2022-01-01", "2022-04-01")))
base_d <- ggplot(df, aes(y = y)) +
  labs(y = NULL) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(colour = "black"),
    panel.grid.minor = element_line(colour = "grey50")
  )

p_maj <- base_d + scale_y_date(date_breaks = "1 month")
p_min <- base_d + scale_y_date(
  date_breaks = "1 month",
  date_minor_breaks = "1 week"
)
p_maj | p_min
```

Because months have varying lengths, minor breaks (spaced exactly 7 days apart) produce slightly uneven spacing within each month.

---

### Labels for Date Scales

Date tick labels are formatted using the `date_labels` argument, which accepts `strptime`-style format strings.

| String | Meaning |
|---|---|
| `%S` | Second (00–59) |
| `%M` | Minute (00–59) |
| `%H` | Hour, 24-hour clock (00–23) |
| `%a` / `%A` | Abbreviated / full weekday name |
| `%d` | Day of month (01–31) |
| `%m` | Month, numeric (01–12) |
| `%b` / `%B` | Abbreviated / full month name |
| `%y` / `%Y` | Year without / with century |

---

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
base_econ <- ggplot(economics, aes(date, psavert)) +
  geom_line(na.rm = TRUE) +
  labs(x = NULL, y = NULL)

p_4digit <- base_econ + scale_x_date(date_breaks = "5 years")
p_2digit <- base_econ + scale_x_date(date_breaks = "5 years", date_labels = "%y")
p_4digit / p_2digit
```

---

The newline character `\n` can be embedded in the format string to stack label components across multiple lines, reducing horizontal crowding:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
lim <- as.Date(c("2004-01-01", "2005-01-01"))

p_compact  <- base_econ + scale_x_date(limits = lim, date_labels = "%b %y")
p_stacked  <- base_econ + scale_x_date(limits = lim, date_labels = "%B\n%Y")
p_compact | p_stacked
```

---

An alternative to `date_labels` is to pass a labelling function to the `labels` argument. The `scales::label_date_short()` function is particularly useful: it automatically abbreviates labels so that repeated information (such as the year) is printed only once per sequence, reducing visual clutter:

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
lim <- as.Date(c("2004-01-01", "2005-01-01"))
base_econ +
  scale_x_date(
    limits = lim,
    labels = scales::label_date_short()
  )
```

# Discrete Position Scales

## Discrete Position Scales

When a categorical (factor or character) variable is mapped to x or y, ggplot2 selects `scale_x_discrete()` or `scale_y_discrete()` automatically. Internally, each category is assigned an integer position, which allows all width-related parameters (bar width, jitter height, etc.) to be specified as proportions of the per-category unit interval.

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
# Explicit integer mapping exposed via annotation
ggplot(mpg, aes(x = hwy, y = class)) +
  geom_point() +
  annotate("text", x = 5, y = 1:7, label = 1:7)
```

---

Because each category occupies unit width on the discrete scale, setting `height = .25` in `geom_jitter()` produces a jitter spanning ±0.25 of the category bin:

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
ggplot(mpg, aes(x = hwy, y = class)) +
  geom_jitter(width = 0, height = .25) +
  annotate("text", x = 5, y = 1:7, label = 1:7)
```

The same logic governs box widths in `geom_boxplot()`: `width = .4` reserves 40% of the category slot for the box.

---

### Limits, Breaks, and Labels

For discrete scales, `limits` defines the **complete ordered set of allowable values** (not endpoints). Values present in the data but absent from `limits` are silently dropped; values in `limits` but absent from the data create empty locations.

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
base_disc <- ggplot(toy, aes(const, txt)) +
  geom_label(aes(label = txt)) +
  scale_x_continuous(breaks = NULL) +
  labs(x = NULL, y = NULL)

p_extra  <- base_disc + scale_y_discrete(limits = c("a", "b", "c", "d", "e"))
p_reord  <- base_disc + scale_y_discrete(limits = c("d", "c", "a", "b"))
base_disc | p_extra | p_reord
```

---

The `breaks` argument selects which values receive axis labels; `labels` maps specific values to custom display strings using a named character vector—without disturbing the order or break set:

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
p_sel_brk <- base_disc + scale_y_discrete(breaks = c("b", "c"))
p_ren_lbl <- base_disc + scale_y_discrete(labels = c(c = "carrot", b = "banana"))
p_sel_brk | p_ren_lbl
```

The function `scales::label_wrap()` is useful for long category names—it automatically wraps text onto multiple lines within a specified character width.

---

### Label Positions

Overlapping axis labels are a common problem with categorical data. The `guides()` function and `guide_axis()` give precise control over label placement:

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
base_cat <- ggplot(mpg, aes(manufacturer, hwy)) + geom_boxplot()
base_cat
```

---

Two effective remedies are **dodging** (staggering labels across multiple rows) and **rotation**:

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
p_dodge  <- base_cat + guides(x = guide_axis(n.dodge = 3))
p_rotate <- base_cat + guides(x = guide_axis(angle = 90))
p_dodge / p_rotate
```

---

`guides()` is a shorthand: the equivalent long form passes `guide` directly inside the scale function.

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
# Equivalent long form
a <- base_cat + scale_x_discrete(guide = guide_axis(n.dodge = 3))
b <- base_cat + scale_x_discrete(guide = guide_axis(angle = 90))
a | b
```

# Binned Position Scales

## Binned Position Scales

Binned position scales occupy the space between continuous and discrete scales. They slice a continuous variable into equally-spaced bins and then treat each bin as a discrete category. The primary function is `scale_x_binned()` (and `scale_y_binned()`).

The most familiar application is histogram-like plots. The example below demonstrates how `geom_bar()` combined with `scale_x_binned()` can approximate `geom_histogram()`:

---

```{r, echo=TRUE, warning=FALSE, out.width="60%", fig.align='center'}
p_hist <- ggplot(mpg, aes(hwy)) + geom_histogram(bins = 8)
p_bin  <- ggplot(mpg, aes(hwy)) + geom_bar()  + scale_x_binned()
p_hist | p_bin
```

---

A more practically useful application pairs `scale_x_binned()` with `geom_count()`. Without binning, `geom_count()` on continuous data produces an over-crowded plot because every distinct coordinate pair gets its own dot:

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
ggplot(mpg, aes(displ, hwy)) + geom_count()
```

---

Adding binned scales on both axes groups nearby points into discrete cells, yielding a much more readable summary:

```{r, echo=TRUE, warning=FALSE, out.width="65%", fig.align='center'}
ggplot(mpg, aes(displ, hwy)) +
  geom_count() +
  scale_x_binned(n.breaks = 15) +
  scale_y_binned(n.breaks = 15)
```

The `n.breaks` argument controls the number of cut points. Binned scales for non-position aesthetics (colour, size, etc.) are discussed in §11.4 and §12.1.2.

---

**Exercises**

1. Using `scale_x_continuous()`, limit the x-axis of a `ggplot(mpg, aes(displ, hwy)) + geom_point()` plot to the range 2–5 in two ways: (a) via `limits`, and (b) via `coord_cartesian()`. Describe the difference in behaviour.

2. Create a time-series line plot of `economics$psavert` against `date`. Set major breaks every 10 years and minor breaks every 2 years. Format the labels to show only the two-digit year.

3. Using the `mpg` dataset, produce a boxplot of `hwy` by `manufacturer`. Apply `guide_axis(angle = 45)` to make the x-axis labels readable.

4. Why does `scale_x_log10()` produce different axis tick labels from `aes(x = log10(displ))`? Under what circumstances would you prefer each approach?

5. Use `scale_x_binned()` with `geom_count()` to visualise the joint distribution of `cty` and `hwy` in the `mpg` dataset. Experiment with different values of `n.breaks`.

## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from [**ggplot2: Elegant Graphics for Data Analysis (3e)**](https://ggplot2-book.org/scales-position.html). All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.

## References
