---
title: "Data Analysis"
subtitle: "Chapter 3: Individual geoms"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "`ggplot2`: Elegant Graphics for Data Analysis"
logo: "logo.svg"
editor: source
bibliography: bib.bib
suppress-bibliography: true
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch3-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

# Basic plot types

## Basic plot types

These geoms are the fundamental building blocks of `ggplot2`. They are useful in their own right, but are also used to construct more complex geoms. Most of these geoms are associated with a named plot: when that geom is used by itself in a plot, that plot has a special name.

Each of these geoms is two dimensional and requires both `x` and `y` aesthetics. All of them understand `colour` (or `color`) and `size` aesthetics, and the filled geoms (bar, tile and polygon) also understand `fill`.

------------------------------------------------------------------------

- `geom_area()` draws an **area plot**, which is a line plot filled to the y-axis (filled lines). Multiple groups will be stacked on top of each other.

- `geom_bar(stat = "identity")` makes a **bar chart**. We need `stat = "identity"` because the default stat automatically counts values (so is essentially a 1d geom, see Section 5.4). The identity stat leaves the data unchanged. Multiple bars in the same location will be stacked on top of one another.

- `geom_line()` makes a **line plot**. The group aesthetic determines which observations are connected; see Chapter 4 for more detail. `geom_line()` connects points from left to right; `geom_path()` is similar but connects points in the order they appear in the data. Both `geom_line()` and `geom_path()` also understand the aesthetic linetype, which maps a categorical variable to solid, dotted and dashed lines.

- `geom_point()` produces a **scatterplot**. `geom_point()` also understands the shape aesthetic.

-----

- `geom_polygon()` draws **polygons**, which are filled paths. Each vertex of the polygon requires a separate row in the data. It is often useful to merge a data frame of polygon coordinates with the data just prior to plotting. Chapter 6 illustrates this concept in more detail for map data.

- `geom_rect()`, `geom_tile()` and `geom_raster()` draw rectangles. `geom_rect()` is parameterised by the four corners of the rectangle, `xmin`, `ymin`, `xmax` and `ymax`. `geom_tile()` is exactly the same, but parameterised by the center of the rect and its size, `x`, `y`, width and height. `geom_raster()` is a fast special case of geom_tile() used when all the tiles are the same size.

- `geom_text()` adds text to a plot. It requires a `label` aesthetic that provides the text to display, and has a number of parameters (`angle`, `family`, `fontface`, `hjust` and `vjust`) that control the appearance of the text.

----------

**Prepare the data**

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
library(ggplot2)
library(patchwork)
df <- data.frame(
  x = c(3, 1, 5), 
  y = c(2, 4, 6), 
  label = c("a","b","c")
)
p <- ggplot(df, aes(x, y, label = label)) + 
  labs(x = NULL, y = NULL) + # Hide axis label
  theme(plot.title = element_text(size = 12)) # Shrink plot title
```

----------

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
p1 <- p + geom_point() + ggtitle("point")
p2 <- p + geom_text() + ggtitle("text")
p3 <- p + geom_bar(stat = "identity", fill = "skyblue") + ggtitle("bar")
p4 <- p + geom_tile(width = 0.5, height = 0.5, fill = "orange") + ggtitle("raster")
(p1 | p2 | p3 | p4)
```

------

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
p5 <- p + geom_line() + ggtitle("line")
p6 <- p + geom_area() + ggtitle("area")
p7 <- p + geom_path() + ggtitle("path")
p8 <- p + geom_polygon() + ggtitle("polygon")
(p5 | p6 | p7 | p8)
```

--------------

**Exercises**

1. What geoms would you use to draw each of the following named plots?

- Scatterplot

- Line chart

- Histogram

- Bar chart

- Pie chart

2. What’s the difference between `geom_path()` and `geom_polygon()`? What’s the difference between `geom_path()` and `geom_line()`?

3. What low-level geoms are used to draw `geom_smooth()`? What about `geom_boxplot()` and `geom_violin()`?



## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from [**ggplot2: Elegant Graphics for Data Analysis (3e)**](https://ggplot2-book.org/getting-started.html). All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.

