---
title: "Data Analysis"
subtitle: "Chapter 16: Faceting"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "`ggplot2`: Elegant Graphics for Data Analysis"
logo: "logo.svg"
editor: source
bibliography: bib.bib
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch16-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

# Faceting

## Overview

**Faceting** generates *small multiples*: a grid of panels, each showing a different subset of the data. Small multiples are a powerful tool for exploratory analysis — you can rapidly compare patterns across subsets and judge whether they are the same or different.

There are three types of faceting:

- `facet_null()` — a single plot (the default)
- `facet_wrap()` — wraps a 1D ribbon of panels into 2D
- `facet_grid()` — a 2D grid of panels defined by row and column variables

Throughout we use a trimmed subset of `mpg` with a manageable number of levels:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
mpg2 <- subset(mpg, cyl != 5 & drv %in% c("4", "f") & class != "2seater")
```

# Facet Wrap

## Facet Wrap

`facet_wrap()` makes a long ribbon of panels and wraps it into 2D — ideal for a single variable with many levels. You control the layout with:

- `ncol` / `nrow` — number of columns or rows (set only one)
- `as.table` — `TRUE` lays out like a table (highest values bottom-right); `FALSE` like a plot (highest top-right)
- `dir` — wrap direction: **h**orizontal or **v**ertical

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
base <- ggplot(mpg2, aes(displ, hwy)) + geom_blank() + xlab(NULL) + ylab(NULL)
a <- base + facet_wrap(~class, ncol = 3)
b <- base + facet_wrap(~class, ncol = 3, as.table = FALSE)
```

---

The same facets, wrapped by rows and in vertical direction:

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
a <- base + facet_wrap(~class, nrow = 3)
b <- base + facet_wrap(~class, nrow = 3, dir = "v")
```

----

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
library(patchwork)
(a|b)
```

# Facet Grid

## Facet Grid

`facet_grid()` lays out panels in a 2D grid defined by a **formula**:

- `. ~ a` — spread `a` across **columns** (aligns vertical scales → compare y position)
- `b ~ .` — spread `b` down **rows** (aligns horizontal scales → compare x position / distributions)
- `b ~ a` — `a` across columns, `b` down rows

```{r, echo=TRUE, warning=FALSE, out.width="32%", fig.align='center'}
a <- base + facet_grid(. ~ cyl)
b <- base + facet_grid(drv ~ .)
c <- base + facet_grid(drv ~ cyl)
```

----

```{r, echo=TRUE, warning=FALSE, out.width="32%", fig.align='center'}
library(patchwork)
(a|b|c)
```

Put the variable with the most levels in the columns to exploit a wide screen.

---

You can use **multiple** variables on a side by adding them: `a + b ~ c + d`. Variables grouped on the same side are **nested** — only combinations present in the data appear. Variables split across rows and columns are **crossed** — all combinations appear, possibly producing empty panels.

# Controlling Scales

## Free vs. Fixed Scales

By default every panel shares the same position scales. The `scales` argument relaxes this:

| Value | Effect |
|---|---|
| `"fixed"` | x and y fixed across all panels (default) |
| `"free_x"` | x free, y fixed |
| `"free_y"` | y free, x fixed |
| `"free"` | both x and y vary |

Fixed scales make it easier to compare *across* panels; free scales make it easier to see patterns *within* a panel. `facet_grid()` adds a constraint: a column shares one x scale and a row shares one y scale.

---

```{r, echo=TRUE, warning=FALSE, out.width="42%", fig.align='center'}
p <- ggplot(mpg2, aes(cty, hwy)) +
  geom_abline() + geom_jitter(width = 0.1, height = 0.1)
a <- p + facet_wrap(~cyl)
b <- p + facet_wrap(~cyl, scales = "free")
```

---

```{r, echo=TRUE, warning=FALSE, out.width="32%", fig.align='center'}
library(patchwork)
(a|b)
```

Free scales are also useful for stacking time series measured on different scales — convert to long form, then `facet_wrap(~variable, scales = "free_y", ncol = 1)`.

---

### The `space` Argument

`facet_grid()` also has a `space` argument (taking the same values as `scales`). When `space = "free"`, each row or column is sized **proportionally** to the range of its scale, so 1 cm maps to the same data range everywhere. This is most useful for categorical scales with differing numbers of levels:

```{r, echo=TRUE, warning=FALSE, out.width="45%", fig.align='center'}
mpg2$model        <- reorder(mpg2$model, mpg2$cty)
mpg2$manufacturer <- reorder(mpg2$manufacturer, -mpg2$cty)
ggplot(mpg2, aes(cty, model)) +
  geom_point() +
  facet_grid(manufacturer ~ ., scales = "free", space = "free") +
  theme(strip.text.y = element_text(angle = 0))
```

# Missing Variables & Grouping

## Missing Faceting Variables

What if a layer's dataset lacks the faceting variable? This commonly arises when adding **context** that should appear in every panel. ggplot2 does the sensible thing: a missing faceting variable is treated as having *all* values, so the layer is drawn in **every** facet.

```{r, echo=TRUE, warning=FALSE, out.width="48%", fig.align='center'}
library(ggplot2)
df1 <- data.frame(x = 1:3, y = 1:3, gender = c("f", "f", "m"))
df2 <- data.frame(x = 2, y = 2)
ggplot(df1, aes(x, y)) +
  geom_point(data = df2, colour = "red", size = 2) +
  geom_point() +
  facet_wrap(~gender)
```

The single red point from `df2` appears in both panels.

## Grouping vs. Faceting

Faceting is an alternative to using aesthetics (colour, shape, size) to distinguish groups. The trade-off is about *position*:

- **Grouping** — groups are close together and may overlap; small differences are easy to see.
- **Faceting** — each group sits in its own panel with no overlap; good when groups overlap heavily, but small differences are harder to spot.

```{r, echo=TRUE, warning=FALSE, out.width="40%", fig.align='center'}
set.seed(1)
df <- data.frame(x = rnorm(120, c(0, 2, 4)), y = rnorm(120, c(1, 2, 1)),
                 z = letters[1:3])
a <- ggplot(df, aes(x, y)) + geom_point(aes(colour = z))   # grouping
b <- ggplot(df, aes(x, y)) + geom_point() + facet_wrap(~z) # faceting
```

---

```{r, echo=TRUE, warning=FALSE, out.width="40%", fig.align='center'}
library(patchwork)
(a|b)
```

---

Faceted comparisons benefit from **annotation**. A useful trick is to put all the data in the background of every panel in grey, so each subset is seen against the whole:

```{r, echo=TRUE, warning=FALSE, out.width="55%", fig.align='center'}
df2 <- dplyr::select(df, -z)
ggplot(df, aes(x, y)) +
  geom_point(data = df2, colour = "grey70") +
  geom_point(aes(colour = z)) +
  facet_wrap(~z)
```

# Continuous Variables

## Faceting Continuous Variables

To facet a continuous variable you must first **discretise** it. ggplot2 provides three helpers:

| Helper | Behaviour |
|---|---|
| `cut_interval(x, n)` | `n` bins of equal *length* |
| `cut_width(x, width)` | bins of equal *width* |
| `cut_number(x, n)` | `n` bins of equal *count* |

```{r, echo=TRUE, warning=FALSE, out.width="32%", fig.align='center'}
mpg2$disp_w <- cut_width(mpg2$displ, 1)
mpg2$disp_i <- cut_interval(mpg2$displ, 6)
mpg2$disp_n <- cut_number(mpg2$displ, 6)
a <- plot <- ggplot(mpg2, aes(cty, hwy)) + geom_point() + labs(x = NULL, y = NULL)
b <- plot + facet_wrap(~disp_w, nrow = 1)
c <- plot + facet_wrap(~disp_i, nrow = 1)
d <- plot + facet_wrap(~disp_n, nrow = 1)
```

----

```{r, echo=TRUE, warning=FALSE, out.width="32%", fig.align='center'}
library(patchwork)
(a|b)/(c|d)
```

Note: the faceting formula does **not** evaluate functions, so create the discretised variable first.

---

**Exercises**

1. Display the distribution of diamond `price` conditional on `cut` and `carat`. Try faceting by `cut` and grouping by `carat`, then the reverse. Which do you prefer, and why?

2. Compare the relationship between `price` and `carat` for each diamond `color`. What makes the groups hard to compare? Is grouping or faceting better here?

3. Why is `facet_wrap()` generally more useful than `facet_grid()`?

---

4. Recreate a plot that facets `mpg2` by `class` while overlaying a single smooth curve fit to the *full* dataset. (Hint: give the smooth layer a dataset without the faceting variable.)

5. Use `facet_grid()` with `space = "free"` to plot `cty` against a reordered `model`, grouped by `manufacturer`. Why does proportional spacing help here?

6. Discretise `displ` three ways with `cut_width()`, `cut_interval()`, and `cut_number()`, then facet by each. How do the panel contents differ?

7. Demonstrate the "all data in grey background" annotation trick on a faceted plot. Why does it aid comparison across panels?

## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from [**ggplot2: Elegant Graphics for Data Analysis (3e)**](https://ggplot2-book.org/facet.html). All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.

## References
