---
title: "Data Analysis"
subtitle: "Chapter 8: Annotations"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "`ggplot2`: Elegant Graphics for Data Analysis"
logo: "logo.svg"
editor: source
bibliography: bib.bib
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch8-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

# Annotations

## Annotations

When building a data visualisation, supplying the plot itself is rarely sufficient — readers also need **context**. Annotations provide that context: they are metadata attached to the visual display that help readers interpret what they are seeing.

Conceptually, annotations are simply data about data. Because of this, ggplot2 implements annotations using the exact same geoms used for plotting data, with a small set of helper functions layered on top for convenience.

------------------------------------------------------------------------

**Chapter roadmap**

- **Section 8.1** — Setting plot and axis titles with `labs()`.
- **Section 8.2** — Adding text labels with `geom_text()` and `geom_label()`.
- **Section 8.3** — Building custom annotations using geoms and `annotate()`.
- **Section 8.4** — Direct labelling with `directlabels`, `ggforce`, and `gghighlight`.
- **Section 8.5** — Annotation across facets.

# Plot and axis titles

## Plot and axis titles

The `labs()` helper function provides a unified interface for setting all text labels associated with a plot: the main title, subtitle, caption, axis labels, and legend titles. Labels are specified as named arguments using the aesthetic or label name as the key:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
library(ggplot2)
library(patchwork)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = factor(cyl))) +
  labs(
    x        = "Engine displacement (litres)",
    y        = "Highway miles per gallon",
    colour   = "Number of cylinders",
    title    = "Mileage by engine size and cylinders",
    subtitle = "Source: https://fueleconomy.gov"
  )
```

------------------------------------------------------------------------

**Mathematical expressions in labels**

Label values are usually plain character strings, with `\n` used to insert line breaks. When a label requires a mathematical expression, wrap it in `quote()` and use R's plotmath syntax. The available symbols and operators are documented at `?plotmath`:

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
values <- seq(from = -2, to = 2, by = .01)
df <- data.frame(x = values, y = values ^ 3)

ggplot(df, aes(x, y)) +
  geom_path() +
  labs(y = quote(f(x) == x^3))
```

------------------------------------------------------------------------

**Markdown in labels with `ggtext`**

The `ggtext` package (Wilke 2020) enables a subset of Markdown syntax — italics, bold, inline HTML — to be rendered inside axis and legend titles. To activate Markdown rendering for a specific element, set the corresponding theme element to `ggtext::element_markdown()`:

----

```{r,echo=T,warning=FALSE,out.width="70%",fig.align='center'}
df <- data.frame(x = 1:3, y = 1:3)
base <- ggplot(df, aes(x, y)) +
  geom_point() +
  labs(x = "Axis title with *italics* and **boldface**")

p1 <- base + ggtitle("Default (no markdown)")
p2 <- base +
  theme(axis.title.x = ggtext::element_markdown()) +
  ggtitle("With element_markdown()")

p1 | p2
```

------------------------------------------------------------------------

**Removing axis labels**

Two strategies exist for removing an axis label, and they behave differently:

- `labs(x = "")` — removes the label text but preserves the space it would have occupied.
- `labs(x = NULL)` — removes both the label text and its allocated space, bringing the axis closer to the plot panel.

The relationship between `labs()` and the underlying scale name system is discussed further in Chapter 14.

# Text labels

## Text labels

Adding text directly to a plot is one of the most common annotation techniques. While labelling every data point is rarely advisable, flagging outliers, notable observations, or small groups of interest is highly effective.

The primary geom for text annotation is `geom_text()`, which places a `label` string at a specified (`x`, `y`) position. It exposes more aesthetics than almost any other geom in ggplot2.

------------------------------------------------------------------------

**The `family` aesthetic — font selection**

The `family` aesthetic accepts a font name. Three values are guaranteed to render correctly on all platforms and output devices:

- `"sans"` — a sans-serif font (the default).
- `"serif"` — a serif font.
- `"mono"` — a monospace font.

---

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
df <- data.frame(x = 1, y = 3:1, family = c("sans", "serif", "mono"))
ggplot(df, aes(x, y)) +
  geom_text(aes(label = family, family = family))
```

------------------------------------------------------------------------

**Font rendering across graphics devices**

System fonts beyond the three guaranteed values may not render consistently across all graphics devices (screen vs. file). Two packages address this:

- **showtext** (<https://github.com/yixuan/showtext>) — converts text to polygons, making font rendering device-independent.
- **extrafont** (<https://github.com/wch/extrafont>) — registers installed system fonts so all devices can use a common format.

Each approach has trade-offs; it is worth testing both in your workflow.

------------------------------------------------------------------------

**The `fontface` aesthetic — text style**

`fontface` accepts three values controlling the typographic weight and slant of the text:

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
df <- data.frame(x = 1, y = 3:1, face = c("plain", "bold", "italic"))
ggplot(df, aes(x, y)) +
  geom_text(aes(label = face, fontface = face))
```

------------------------------------------------------------------------

**The `hjust` and `vjust` aesthetics — alignment**

`hjust` (horizontal justification) and `vjust` (vertical justification) control how a label is positioned relative to its anchor point. Both accept named strings or numeric values in [0, 1].

Particularly useful values are `"inward"` / `"outward"`, which align text *towards* or *away from* the centre of the plot regardless of the label's position — ensuring that labels stay within the plot boundaries:

---

```{r,echo=T,warning=FALSE,out.width="70%",fig.align='center'}
df <- data.frame(
  x    = c(1, 1, 2, 2, 1.5),
  y    = c(1, 2, 1, 2, 1.5),
  text = c("bottom-left", "top-left", "bottom-right", "top-right", "center")
)

p1 <- ggplot(df, aes(x, y)) + geom_text(aes(label = text)) +
  ggtitle("Default alignment")
p2 <- ggplot(df, aes(x, y)) +
  geom_text(aes(label = text), vjust = "inward", hjust = "inward") +
  ggtitle("Inward alignment")
p1 | p2
```

------------------------------------------------------------------------

**The `size` aesthetic — font size**

Font size is specified in **millimetres (mm)**, not points. This keeps text sizes consistent with the units used for all other size aesthetics in ggplot2. To convert from points to mm: multiply by 25.4 / 72.27.

**The `angle` parameter — rotation**

`angle` rotates the label by the specified number of degrees, measured counter-clockwise from horizontal.

------------------------------------------------------------------------

**Nudging labels — `nudge_x` and `nudge_y`**

When labelling existing data points, overlapping text and points is a common problem. The `nudge_x` and `nudge_y` parameters shift the label by a fixed amount relative to the data point, without altering the position used for the `x`/`y` aesthetics:

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
df <- data.frame(treatment = c("a", "b", "c"), response = c(1.2, 3.4, 2.5))

ggplot(df, aes(treatment, response)) +
  geom_point() +
  geom_text(
    aes(label = paste0("(", response, ")")),
    nudge_x = -0.3
  ) +
  ylim(1.1, 3.6)
```

------------------------------------------------------------------------

**Removing overlapping labels — `check_overlap`**

Setting `check_overlap = TRUE` automatically suppresses labels that would overlap previously plotted labels. Labels are evaluated in data frame row order — earlier rows take priority. If you sort your data by importance before plotting, the result is a label set that emphasises the most important observations:

```{r,echo=T,warning=FALSE,out.width="70%",fig.align='center'}
p1 <- ggplot(mpg, aes(displ, hwy)) +
  geom_text(aes(label = model)) + xlim(1, 8) +
  ggtitle("check_overlap = FALSE")
p2 <- ggplot(mpg, aes(displ, hwy)) +
  geom_text(aes(label = model), check_overlap = TRUE) + xlim(1, 8) +
  ggtitle("check_overlap = TRUE")
p1 | p2
```

------------------------------------------------------------------------

**`geom_label()` — text with a background box**

`geom_label()` is a variant of `geom_text()` that draws a rounded rectangle behind the text. The opaque background ensures legibility on complex or colourful plot backgrounds:

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
label <- data.frame(
  waiting   = c(55, 80),
  eruptions = c(2, 4.3),
  label     = c("peak one", "peak two")
)

ggplot(faithfuld, aes(waiting, eruptions)) +
  geom_tile(aes(fill = density)) +
  geom_label(data = label, aes(label = label))
```

------------------------------------------------------------------------

**Common challenges with text labels**

- **Labels do not expand plot limits.** Because a label has a fixed physical size (e.g., 3 cm) that is independent of plot dimensions, ggplot2 cannot automatically expand the plot boundaries to accommodate it. Adjust `xlim()` / `ylim()` manually.

- **Avoiding overlap with many labels.** The `ggrepel` package (<https://github.com/slowkow/ggrepel>) by Kamil Slowikowski provides `geom_text_repel()`, which optimises label placement to minimise overlaps:

---

```{r,echo=T,warning=FALSE,message=FALSE,out.width="65%",fig.align='center'}
set.seed(42)
mini_mpg <- mpg[sample(nrow(mpg), 20), ]
ggplot(mpg, aes(displ, hwy)) +
  geom_point(colour = "red") +
  ggrepel::geom_text_repel(data = mini_mpg, aes(label = class))
```

- **Fitting labels inside bars or tiles.** The `ggfittext` package by Claus Wilke provides geoms that automatically resize or trim text to fit within a specified bounding region.

# Building custom annotations

## Building custom annotations

Beyond text labels, ggplot2 provides several geoms suited to annotating specific types of information. These geoms work exactly like data-plotting geoms — they accept either a separate data frame or are placed using the `annotate()` helper function.

**Key annotation geoms:**

- `geom_rect()` — fills a rectangular region; uses `xmin`, `xmax`, `ymin`, `ymax` aesthetics.
- `geom_segment()` / `geom_line()` / `geom_path()` — draw line segments or paths; all accept an `arrow` argument for arrowheads (created with `arrow(angle, length, ends, type)`).
- `geom_curve()` — draws a curved arc between two points; useful for connecting a label to a data point.
- `geom_vline()` / `geom_hline()` / `geom_abline()` — draw reference lines spanning the full extent of the plot.

------------------------------------------------------------------------

**Using `-Inf` and `Inf` as position values**

When a geom should extend to the very edge of the plot panel — regardless of the data range — use `-Inf` (bottom/left limit) and `Inf` (top/right limit) as position values. This is especially useful for `geom_rect()` when creating background shading.

------------------------------------------------------------------------

**Example: annotating a time series with presidential terms**

The `economics` dataset records US unemployment over time. We will overlay shaded rectangles indicating presidential terms, vertical lines at term boundaries, and president name labels, using four separate layers:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
presidential_sub <- subset(presidential, start > economics$date[1])

fig <- ggplot(economics) +
  geom_rect(
    aes(xmin = start, xmax = end, fill = party),
    ymin = -Inf, ymax = Inf, alpha = 0.2,
    data = presidential_sub
  ) +
  geom_vline(
    aes(xintercept = start),
    data = presidential_sub, colour = "grey50", alpha = 0.5
  ) +
  geom_text(
    aes(x = start, y = 2500, label = name),
    data = presidential_sub, size = 3, vjust = 0, hjust = 0, nudge_x = 50
  ) +
  geom_line(aes(date, unemploy)) +
  scale_fill_manual(values = c("blue", "red")) +
  labs(x = "Date", y = "Unemployment")
```

----

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
fig 
```

Each layer draws from its **own data frame** — note how `data = presidential_sub` is specified inside each annotation geom while `geom_line()` inherits the default `economics` data. This is the standard pattern for mixing annotation data with the primary dataset.

------------------------------------------------------------------------

**`annotate()` — a shorthand for single annotations**

Adding a single annotation via a geom requires constructing a one-row data frame, which is verbose. The `annotate()` helper function constructs this data frame internally:

```{r,echo=T,eval=FALSE}
# Verbose approach (using a data frame)
ggplot(economics, aes(date, unemploy)) +
  geom_line() +
  geom_text(
    aes(x, y, label = caption),
    data = data.frame(x = xrng[1], y = yrng[2], caption = caption),
    hjust = 0, vjust = 1, size = 4
  )
```

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
# Concise approach using annotate()
yrng    <- range(economics$unemploy)
xrng    <- range(economics$date)
caption <- paste(strwrap(
  "Unemployment rates in the US have varied a lot over the years", 40
), collapse = "\n")

fig <- ggplot(economics, aes(date, unemploy)) +
  geom_line() +
  annotate(
    geom = "text", x = xrng[1], y = yrng[2],
    label = caption, hjust = 0, vjust = 1, size = 4
  )
```

---

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
fig
```

------------------------------------------------------------------------

**Highlighting a subset with manual annotation**

A common pattern is to highlight a subgroup by plotting it twice — once with a larger, differently-coloured point underneath the main scatter layer — and then labelling it with `annotate()`:

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
p <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(
    data  = dplyr::filter(mpg, manufacturer == "subaru"),
    colour = "orange", size = 3
  ) +
  geom_point()

p +
  annotate(geom = "point", x = 5.5, y = 40, colour = "orange", size = 3) +
  annotate(geom = "point", x = 5.5, y = 40) +
  annotate(geom = "text",  x = 5.6, y = 40, label = "subaru", hjust = "left")
```

------------------------------------------------------------------------

**Curved connectors with `geom_curve()`**

When a label must be placed away from its target to avoid clutter, `geom_curve()` draws a curved arc with an optional arrowhead to visually connect the two. Use `curvature` to control the arc's degree of bending (positive = clockwise, negative = counter-clockwise):

```{r,echo=T,warning=FALSE,out.width="65%",fig.align='center'}
p +
  annotate(
    geom      = "curve",
    x = 4, y = 35, xend = 2.65, yend = 27,
    curvature = .3,
    arrow     = arrow(length = unit(2, "mm"))
  ) +
  annotate(geom = "text", x = 4.1, y = 35, label = "subaru", hjust = "left")
```

# Direct labelling

## Direct labelling

**Direct labelling** refers to the practice of placing group labels directly on the plot — adjacent to or within the data they identify — rather than in a separate legend. This approach typically reduces the visual distance between a label and its data, making the chart faster to read.

Three packages in the broader ggplot2 ecosystem provide tools for automated direct labelling.

------------------------------------------------------------------------

**`directlabels` — smart label placement**

The `directlabels` package (Hocking) provides `geom_dl()`, which adds group labels at algorithmically determined positions. The `method` argument selects a placement strategy:

---

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
p1 <- ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point() +
  ggtitle("Legend (default)")

p2 <- ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point(show.legend = FALSE) +
  directlabels::geom_dl(aes(label = class), method = "smart.grid") +
  ggtitle("Direct labels (smart.grid)")

p1 | p2
```

`"smart.grid"` is a good default for scatterplots. For line plots and frequency polygons, other methods (e.g., `"last.points"`, `"angled.boxes"`) are more appropriate. See the directlabels website (<http://directlabels.r-forge.r-project.org>) for a complete catalogue.

------------------------------------------------------------------------

**`ggforce` — ellipse highlight marks**

The `ggforce` package (Pedersen) extends ggplot2 with a range of specialised geoms, including `geom_mark_ellipse()`. This geom draws an elliptical highlight around each group of points and adds a label, making group membership immediately apparent without a legend:

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  ggforce::geom_mark_ellipse(aes(label = cyl, group = cyl))
```

------------------------------------------------------------------------

**`gghighlight` — emphasising selected data**

The `gghighlight` package (Yutani) is designed to visually emphasise a subset of observations while showing the full dataset in the background for context. It is particularly effective for longitudinal or grouped data:

---

```{r,echo=T,warning=FALSE,message=FALSE,out.width="70%",fig.align='center'}
data(Oxboys, package = "nlme")

ggplot(Oxboys, aes(age, height, group = Subject)) +
  geom_line() +
  geom_point() +
  gghighlight::gghighlight(Subject %in% 1:3)
```

The highlighted subjects are drawn in full colour; all other trajectories are rendered in a muted grey, providing the visual context of the whole dataset without competing for attention.

# Annotation across facets

## Annotation across facets

Annotations become especially powerful in faceted plots, where they can provide a shared visual reference that makes between-panel comparisons easier.

**Motivating example: faceted diamond price vs carat**

The plot below facets the relationship between `log10(carat)` and `log10(price)` by diamond `cut`. The within-facet relationship is clear, but the subtle differences in slope or intercept across cut grades are difficult to perceive without a reference:

---

```{r,echo=T,warning=FALSE,message=FALSE,out.width="80%",fig.align='center'}
ggplot(diamonds, aes(log10(carat), log10(price))) +
  geom_bin2d() +
  facet_wrap(vars(cut), nrow = 1)
```

------------------------------------------------------------------------

**Adding a shared reference line across all facets**

Fitting a single regression line to the full dataset and overlaying it in every facet gives the eye a fixed reference. Deviations from this line — or differences in the cloud's position relative to the line — now pop out visually:

---

```{r,echo=T,warning=FALSE,message=FALSE,out.width="80%",fig.align='center'}
mod_coef <- coef(lm(log10(price) ~ log10(carat), data = diamonds))

ggplot(diamonds, aes(log10(carat), log10(price))) +
  geom_bin2d() +
  geom_abline(
    intercept = mod_coef[1],
    slope     = mod_coef[2],
    colour    = "white",
    linewidth = 1
  ) +
  facet_wrap(vars(cut), nrow = 1)
```

Each facet now shows its data relative to the **same** regression line, making cross-facet comparisons unambiguous. Using a white line against the 2D bin colour scale makes it legible without overpowering the data.

------------------------------------------------------------------------

**`gghighlight` inside facets — greyed background data**

Another effective pattern is to facet by group while showing the full dataset greyed out in every panel. Each facet highlights one group in colour while all other data remains visible for context. `gghighlight::gghighlight()` with no filtering argument achieves this automatically:

---

```{r,echo=T,warning=FALSE,message=FALSE,out.width="75%",fig.align='center'}
ggplot(mpg, aes(displ, hwy, colour = factor(cyl))) +
  geom_point() +
  gghighlight::gghighlight() +
  facet_wrap(vars(cyl))
```

Each panel foregrounds the data for one cylinder group while maintaining the full context of the dataset in grey. This technique elegantly combines the detail of faceting with the contextual richness of showing all data simultaneously.

## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from [**ggplot2: Elegant Graphics for Data Analysis (3e)**](https://ggplot2-book.org/annotations.html). All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.

## References
