## Problem sets

- [In-class Exercise 1](https://yyliou.github.io/da/da-ps/ex1.html)
- [In-class Exercise 2](https://yyliou.github.io/da/da-ps/ex2.html)
- [In-class Exercise 3](https://yyliou.github.io/da/da-ps/ex3.html)
- [In-class Exercise 4](https://yyliou.github.io/da/da-ps/ex4.html)
- [In-class Exercise 5](https://yyliou.github.io/da/da-ps/ex5.html)
- [In-class Exercise 6](https://yyliou.github.io/da/da-ps/ex6.html)
- [In-class Exercise 7](https://yyliou.github.io/da/da-ps/ex7.html)
- [In-class Exercise 8](https://yyliou.github.io/da/da-ps/ex8.html)
- [In-class Exercise 9](https://yyliou.github.io/da/da-ps/ex9.html)
- [In-class Exercise 10](https://yyliou.github.io/da/da-ps/ex10.html)
- [In-class Exercise 11](https://yyliou.github.io/da/da-ps/ex11.html)
- [In-class Exercise 12](https://yyliou.github.io/da/da-ps/ex12.html)

## Submission Requirements

> ⚠️ **All assignments must be submitted using the official template provided below.**
>
> Submissions that are missing any of the following elements will receive a grade of zero. This policy is non-negotiable.

### Template

Copy the following template as the starting point for every submission. Replace all placeholder values before submitting.

````{r,eval=F}
---
title: "Ex. X"
author: "Student ID"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
print(uuid::UUIDgenerate())
```

## Question 1

*(Write your response here.)*

## Question 2

*(Write your response here.)*
````

### Checklist

Before submitting, verify that your document satisfies all of the following criteria:

-   [ ] The `title` field in the YAML header is completed (e.g., `Ex. 1`).
-   [ ] The `author` field includes your student ID *only*.
-   [ ] The `date` field is present and left exactly as it appears in the template.
-   [ ] The UUID code chunk (`print(uuid::UUIDgenerate())`) from the template is present and unmodified.
-   [ ] The final document must be submitted strictly in `.pdf` format.
-   [ ] The document's name should be `exN-id.pdf` (e.g., `ex1-D10627008.pdf`).

## Submit Your Assignments

Upload your PDF [here](https://u.pcloud.link/publink/show?code=kZaVeI5ZjmzQXBbRlv7XxLzlO1xTDB92w9Kk) and place it in the corresponding folder (e.g., `ex1` for Exercise 1).

## Check your grade

- [Here](https://u.pcloud.link/publink/show?code=kZgSqI5ZXpSmLylHANukDaMOW6lwHzVhiAA7)

